---
title: "Working with realized GDS and spatial FDR"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Working with realized GDS and spatial FDR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(fmrigds)
set.seed(1)
```

# Verbs on realized GDS via as_plan()

All verbs accept realized `gds` objects. Internally, fmrigds converts a realized
GDS to a lazy plan using `as_plan(gds)` and uses an in-memory adapter, so you
can chain operations without going back to disk.

```{r}
# Build a tiny ROI CSV
tmp <- tempfile(fileext = ".csv")
writeLines(c(
  "sample,subject,contrast,beta,var",
  "ROI_1,s1,c1,0.5,0.04",
  "ROI_1,s2,c1,0.9,0.04",
  "ROI_2,s1,c1,0.7,0.09",
  "ROI_2,s2,c1,1.2,0.09"
), tmp)

# Plan -> realized GDS
g0 <- compute(gds(tmp))

# Reduce directly on realized object
g_fe <- reduce(g0, method = "meta:fe") |> compute()

# Post-hoc on realized object (apply after a reduce so p-values exist)
g_bh <- reduce(g0, method = "meta:fe") |> posthoc(method = "fdr:bh") |> compute()

names(assays(g_fe))  # group-level assays
names(assays(g_bh))  # includes q
```

# Spatial FDR (fdr:spatial)

`fdr:spatial` performs a group-wise adjustment by:

- Aggregating p-values within groups via Simes’ method
- Applying a weighted BH across groups (weights ~ group size, normalized)
- Broadcasting group q-values back to samples

Requirements:

- Provide `p` (or `z` / `t` + `df`) so p-values can be computed/derived
- Provide `options$group`, a vector of length equal to the number of samples
  (first dimension), defining group membership (e.g., ROI/parcel IDs). `NA`
  entries are ignored. Groups can be numeric, character, or factors.

```{r}
# Create a realized GDS from p-values
gp <- compute(gds(tmp, effect_cols = list(p = "var")))  # reusing column for demo

# Two groups, one sample per group (matches 2 samples in tmp)
grp <- c(1, 2)

# Apply spatial FDR
gq <- posthoc(gp, method = "fdr:spatial", options = list(group = grp)) |> compute()
assay(gq, "q")
```
