---
title: Driving fmrigds from the command line
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Driving fmrigds from the command line}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("albersdown", quietly = TRUE)) {
  if (requireNamespace("systemfonts", quietly = TRUE)) try(albersdown::albers_register_fonts(), silent = TRUE)
  if (requireNamespace("ggplot2", quietly = TRUE)) try(ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset)), silent = TRUE)
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)

library(fmrigds)

if (!exists(".cli_main", envir = asNamespace("fmrigds"), inherits = FALSE) &&
    requireNamespace("pkgload", quietly = TRUE)) {
  pkgload::load_all(".", helpers = FALSE, export_all = FALSE, quiet = TRUE)
}

cli_main <- getFromNamespace(".cli_main", "fmrigds")

cli_run <- function(args) {
  args <- as.character(args)
  paste(capture.output(cli_main(args)), collapse = "\n")
}

cli_text <- function(...) {
  cli_run(unlist(list(...), use.names = FALSE))
}

cli_json <- function(...) {
  jsonlite::fromJSON(cli_text(...))
}

cli_show <- function(...) {
  args <- as.character(unlist(list(...), use.names = FALSE))
  cat("$ fmrigds ", paste(shQuote(args), collapse = " "), "\n", sep = "")
  cat(cli_run(args), "\n", sep = "")
  invisible(NULL)
}

node_table <- function(nodes) {
  if (is.null(nodes)) {
    return(data.frame(op = character(), summary = character()))
  }
  if (is.data.frame(nodes)) {
    keep <- intersect(c("op", "summary"), names(nodes))
    return(nodes[keep])
  }
  data.frame(
    op = vapply(nodes, function(x) x$op, character(1)),
    summary = vapply(nodes, function(x) x$summary, character(1)),
    stringsAsFactors = FALSE
  )
}

stats_path <- tempfile(fileext = ".csv")
writeLines(c(
  "sample,subject,contrast,time,beta,var",
  "ROI_1,sub-01,baseline,0,0.50,0.04",
  "ROI_1,sub-01,task,1,0.80,0.04",
  "ROI_1,sub-02,baseline,0,0.40,0.05",
  "ROI_1,sub-02,task,1,0.90,0.05",
  "ROI_1,sub-03,baseline,0,0.45,0.05",
  "ROI_1,sub-03,task,1,1.05,0.05",
  "ROI_1,sub-04,baseline,0,0.52,0.05",
  "ROI_1,sub-04,task,1,1.12,0.05",
  "ROI_2,sub-01,baseline,0,0.60,0.03",
  "ROI_2,sub-01,task,1,0.70,0.03",
  "ROI_2,sub-02,baseline,0,0.55,0.02",
  "ROI_2,sub-02,task,1,0.95,0.02",
  "ROI_2,sub-03,baseline,0,0.58,0.02",
  "ROI_2,sub-03,task,1,1.10,0.02",
  "ROI_2,sub-04,baseline,0,0.62,0.02",
  "ROI_2,sub-04,task,1,1.18,0.02"
), stats_path)

cov_path <- tempfile(fileext = ".csv")
writeLines(c(
  "subject,age,group",
  "sub-01,29,control",
  "sub-02,33,control",
  "sub-03,37,patient",
  "sub-04,41,patient"
), cov_path)

quick_out <- tempfile(fileext = ".csv")
plan_path <- tempfile(fileext = ".json")
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

## Why use the CLI?

Sometimes you want the `fmrigds` pipeline without opening an R session for each
run. The CLI gives you the same lazy plan grammar that you use in R, but in a
form that fits scripts, shell pipelines, and scheduled jobs.

The common path is short: inspect a source, save a plan, preview a small block,
then execute it. When you need more control, the same CLI also accepts metadata
tables, model formulas, reducer options, post-hoc options, and saved plan JSON.

At the terminal, the workflow looks like this:

```bash
fmrigds probe --input group.csv
fmrigds plan --input group.csv --reduce fixed --save-plan plan.json
fmrigds preview --load-plan plan.json --n 3
fmrigds run --load-plan plan.json --out results.h5
```

The chunks below call the same CLI parser from inside R so the examples stay
runnable in the vignette.

## What are we driving?

We'll use a small long-table input with ROI labels, four subjects, and two
contrasts. The extra `time` column will become contrast metadata later.

```{r cli-input}
read.csv(stats_path, stringsAsFactors = FALSE)
```

## How do you get a result quickly?

If you already know the source and the reducer you want, `run` is the shortest
path from input table to output file.

```{r cli-quick-win}
cli_show(
  "run",
  "--input", stats_path,
  "--subset", "contrast=task",
  "--reduce", "fixed",
  "--out", quick_out
)
```

```{r cli-quick-win-output}
head(read.csv(quick_out, stringsAsFactors = FALSE))
```

```{r cli-quick-win-check, include = FALSE}
quick_df <- read.csv(quick_out, stringsAsFactors = FALSE)
stopifnot(file.exists(quick_out))
stopifnot(nrow(quick_df) == 2L)
stopifnot(all(c("sample", "subject", "contrast", "beta", "var") %in% names(quick_df)))
stopifnot(all(is.finite(quick_df$beta)))
```

That is the "easy path": subset to the contrast you want, reduce across
subjects, and write a tidy result. For one-off jobs or shell scripts, that is
often enough.

## How do you inspect a source before you commit to a model?

`probe` tells you what the CLI sees before anything is computed: the adapter,
the dimensions, the assays, and whether the source has enough structure for the
next step.

```{r cli-probe}
probe <- cli_json(
  "probe",
  "--input", stats_path,
  "--contrast-data-cols", "time",
  "--json"
)

data.frame(
  adapter = probe$adapter,
  samples = probe$dims[1],
  subjects = probe$dims[2],
  contrasts = probe$dims[3],
  assays = paste(probe$assays, collapse = ", "),
  has_contrast_data = probe$has_contrast_data,
  stringsAsFactors = FALSE
)
```

```{r cli-probe-check, include = FALSE}
stopifnot(identical(probe$adapter, "tabular"))
stopifnot(identical(as.integer(unlist(probe$dims)), c(2L, 4L, 2L)))
stopifnot(isTRUE(probe$has_contrast_data))
```

Here `--contrast-data-cols time` promotes the `time` column into
contrast-level metadata. That matters when you move from a simple group summary
to a model that needs contrast annotations.

## How do you separate planning from execution?

`plan` is useful when you want to describe the workflow once, save it, review
it, and execute it later.

```{r cli-plan}
plan_info <- cli_json(
  "plan",
  "--input", stats_path,
  "--contrast-data-cols", "time",
  "--subset", "contrast=task",
  "--reduce", "fixed",
  "--posthoc", "fdr:bh",
  "--save-plan", plan_path,
  "--json"
)

knitr::kable(node_table(plan_info$nodes), row.names = FALSE)
```

```{r cli-plan-check, include = FALSE}
loaded_plan <- load_plan(plan_path)
stopifnot(file.exists(plan_path))
stopifnot(identical(vapply(loaded_plan$nodes, `[[`, character(1), "op"),
                    c("subset_axis", "reduce", "posthoc")))
stopifnot(identical(rownames(contrast_data(loaded_plan)), c("baseline", "task")))
stopifnot(all(contrast_data(loaded_plan)$time == c(0, 1)))
```

This is the point where the CLI stops being a thin wrapper around `compute()`
and becomes a real planning tool. You can keep the saved JSON under version
control, review the steps, and reuse the same plan across runs.

## How do you preview a saved plan?

`preview` runs a small sample block through the current plan. That is useful
when you want to check the output contract without writing the full result.

```{r cli-preview}
preview_out <- cli_json(
  "preview",
  "--load-plan", plan_path,
  "--n", "1",
  "--show-assay", "beta,q",
  "--json"
)

preview_out$data
```

```{r cli-preview-check, include = FALSE}
stopifnot(nrow(preview_out$data) == 1L)
stopifnot(identical(preview_out$data$sample, "ROI_1"))
stopifnot(identical(preview_out$data$subject, "meta"))
stopifnot(identical(preview_out$data$contrast, "task"))
stopifnot(all(is.finite(preview_out$data$beta)))
stopifnot(all(preview_out$data$q >= 0 & preview_out$data$q <= 1))
```

The preview already includes the derived and post-hoc assays from the saved
plan. In practice, this is the fastest way to confirm that a workflow is
producing the fields you expect before you launch a longer batch run.

## How do you reach more complex workflows?

The CLI is not limited to fixed-effects summaries. You can also attach
subject-level covariates, specify formulas, and hand reducer-specific options
to more complex models.

```{r cli-meta-regression}
cli_show(
  "preview",
  "--input", stats_path,
  "--subset", "contrast=task",
  "--col-data", cov_path,
  "--reduce", "meta:fe_reg",
  "--formula", "~ age",
  "--show-assay", "coef:age",
  "--n", "2"
)
```

```{r cli-meta-regression-check, include = FALSE}
reg_preview <- cli_json(
  "preview",
  "--input", stats_path,
  "--subset", "contrast=task",
  "--col-data", cov_path,
  "--reduce", "meta:fe_reg",
  "--formula", "~ age",
  "--show-assay", "coef:age",
  "--n", "2",
  "--json"
)
stopifnot("coef:age" %in% names(reg_preview$data))
stopifnot(all(is.finite(reg_preview$data[["coef:age"]])))
```

The same pattern extends to other advanced switches:

- `--source-option key=value` passes adapter-specific arguments through `gds()`.
- `--reduce-option key=value` passes reducer options through `reduce()`.
- `--posthoc-option key=value` passes method-specific options through `posthoc()`.
- `--write-option key=value` controls exporters when you write to disk.
- `--load-plan` and `--save-plan` let you mix shell-driven execution with
  plan-first workflows.

## How do you discover available methods?

`list` exposes the registries that the CLI can target. You can use it to see
which reducers and post-hoc methods are available in the current installation.

```{r cli-list}
reducers <- cli_json("list", "reducers", "--json")

head(
  data.frame(
    name = reducers$name,
    requires = vapply(reducers$requires, paste, collapse = ", ", character(1)),
    provides = vapply(reducers$provides, paste, collapse = ", ", character(1)),
    stringsAsFactors = FALSE
  ),
  6
)
```

If you are scripting around the CLI, this registry view is a good way to make
your wrapper validate method names before a larger run.

## Where should you go next?

Once you are comfortable with the CLI surface, the R vignettes fill in the
modeling details:

- `vignette("fmrigds")` for the core lazy workflow
- `vignette("spatial-operations")` for mapping, masking, and alignment
- `vignette("as-plan-and-spatial-fdr")` for post-hoc and spatial FDR workflows

The CLI is best when you want a reproducible command surface. When you need
custom transforms or deeply programmatic branching, save a plan from the CLI
and continue in R with `load_plan()`, `explain_plan()`, and `compute()`.
