---
title: Reading native HDF5 and legacy fmristore files
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Reading native HDF5 and legacy fmristore files}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("albersdown", quietly = TRUE)) {
  if (requireNamespace("systemfonts", quietly = TRUE)) try(albersdown::albers_register_fonts(), silent = TRUE)
  if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  eval = FALSE
)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```


## When do you need this?

HDF5 is a container that can hold several different schemas. fmrigds uses a
current native `/gds` schema for files written by `write_out()`, and it also
reads older schemas created by the fmristore package. Those legacy files can
contain voxelwise statistical maps (labeled volumes), parcel-level timeseries,
or latent-space decompositions. fmrigds auto-detects the layout and converts
the supported structure to a GDS.

All examples below use `eval = FALSE` because they require HDF5 files that are
not shipped with the package. To run them locally, you need `hdf5r`:

```r
install.packages("hdf5r")
```

## Two paths for reading data

fmrigds supports two kinds of HDF5 input. The path you use depends on how the
file was created:

**Current path (recommended)**: Native fmrigds `/gds` format
:   Files written by `write_out(..., format = "h5")` contain a `/gds` group
    containing assays, axes, space, axis tables, metadata, and provenance
    records.

**Compatibility path (legacy)**: Original fmristore layouts
:   Files created by the fmristore package use one of three layout types:
    LabeledVolumeSet, Parcellated, or LatentNeuroVec. fmrigds auto-detects
    the layout and converts to GDS.

If you control file creation, use the current native path. If you are reading
an existing fmristore file, the compatibility path works automatically with
`gds(path)`. The word *legacy* refers to those older fmristore schemas, not to
HDF5 generally and not to the native `/gds` format.

## Native fmrigds `/gds` layout

The simplest case is a file with a `/gds` group. The current reader restores
assays, axes, spatial structure, axis tables, and supported registered map
families:

```r
library(fmrigds)

plan <- gds("group_store.h5")
g <- compute(plan)

names(assays(g))
space(g)
col_data(g)
row_data(g)
contrast_data(g)
```

Create files in this format with `write_out()`:

```r
gds(source) |>
  reduce(method = "fixed") |>
  write_out("group_results.h5", format = "h5") |>
  compute()
```

## Labeled volumes (LabeledVolumeSet)

The most common fmristore layout for first-level outputs. It stores
statistical maps in voxel space with one dataset per contrast.

Structure inside the HDF5 file:

```
/mask         3D binary mask (which voxels have data)
/data/
  cope1       Packed vector of values for masked voxels
  cope2       ...
/labels       List of contrast names
```

```r
plan <- gds("subject01_lvs.h5")
g <- compute(plan)

space(g)               # voxel space
names(assays(g))       # "beta", "var", etc.
contrasts(g)           # e.g., c("cope1", "cope2")
dim(assay(g, "beta"))  # [n_voxels x n_subjects x n_contrasts]
```

Contrast names containing special characters are sanitized
(e.g., `"cope-1"` becomes `"cope_1"`).

## Parcellated data (H5ParcellatedScan)

Parcellated layouts store timeseries data for brain parcels (ROIs). Each file
contains:

```
/cluster_map    3D integer array (voxels -> parcel IDs)
/scans/
  001           Matrix: parcels x timepoints
  002           ...
```

The challenge: fmristore stores *timeseries*, but GDS needs *contrasts*. You
must tell fmrigds how to convert timepoints into contrasts using a
**temporal policy**.

### Temporal policies

**`"as_is"` (default)** --- treat each timepoint as a separate contrast:

```r
plan <- gds("parcellated.h5")
g <- compute(plan)
contrasts(g)  # c("t_001", "t_002", "t_003", ...)
```

Use this when you want to analyse the full timecourse.

**`"mean"`** --- average all timepoints into one contrast:

```r
plan <- gds("parcellated.h5", temporal_policy = "mean")
g <- compute(plan)
contrasts(g)  # "mean"
```

Only `beta` is averaged; `var`/`se` become `NA` (no uncertainty is
fabricated).

**`"design"`** --- apply a contrast matrix to extract specific effects:

```r
# Extract two contrasts from 3 timepoints
C <- cbind(early = c(1, 0, 0), late = c(0, 1, 0))

plan <- gds(
  "parcellated.h5",
  temporal_policy = "design",
  contrast_matrix = C,
  contrast_names = c("early", "late")
)

g <- compute(plan)
contrasts(g)           # c("early", "late")
dim(assay(g, "beta"))  # [n_parcels x n_subjects x 2]
```

Variance propagation: if the source file includes `/var`, fmrigds applies
`Var(C'y) = (C^2)' Var(y)` to propagate uncertainty through the linear
combination.

### Space type

After reading, the GDS has parcel-type space:

```r
g <- compute(gds("parcellated.h5"))
space(g)  # parcels space with labels
```

## Latent bases (LatentNeuroVec)

Latent layouts store dimensionality-reduced data where brain activity is
represented by a small number of spatial components (e.g., from PCA or ICA).

Structure:

```
/basis          Matrix: components x voxels
/scans/
  001           Embeddings: how much each component is expressed over time
  002           ...
```

```r
plan <- gds("latent_small.h5")
g <- compute(plan)

space(g)               # basis space
dim(assay(g, "beta"))  # [k_components x n_subjects x n_contrasts]
```

Each "sample" is a spatial component (a brain network). If the original data
had no uncertainty estimates, `var`/`se` will be `NA`.

## Reading subsets efficiently

For large datasets, read specific slices using the `block` parameter instead
of loading everything into memory:

```r
# Read only first 10 samples and first contrast
raw <- compute(
  plan,
  assays = "beta",
  block = list(sample = 1:10, contrast = 1)
)

# Read specific subjects
raw <- compute(
  plan,
  assays = c("beta", "var"),
  block = list(subject = c("sub-01", "sub-03"))
)
```

## Provenance tracking

Every `compute()` operation records what was done:

```r
plan <- gds("data.h5") |>
  reduce(method = "fixed")

g <- compute(plan)
metadata(g)$provenance$log
```

The native writer serializes GDS metadata and writes the provenance log into
the archive. The current reader does not promise to reconstruct every
arbitrary custom metadata field across versions. Treat application-specific
metadata as an explicit verification point after reading; the core GDS data
model listed above is the supported read contract.

## Quick reference

| Scenario | What to do |
|---|---|
| Creating new files | `write_out(..., format = "h5")` (native `/gds`) |
| Existing fmristore files | `gds(path)` --- layout auto-detected |
| Parcellated timeseries | Set `temporal_policy` (`"as_is"`, `"mean"`, or `"design"`) |
| Large files | Use `block = list(...)` to read subsets |
| Reproducibility | Native archives store metadata and provenance records; verify custom fields after reading |

## Next steps

- `vignette("fmrigds")` --- getting started with the core pipeline
- `vignette("spatial-operations")` --- masking, alignment, and space
  transformations
- `vignette("as-plan-and-spatial-fdr")` --- post-hoc corrections and spatial
  FDR
- `?gds` for the full list of adapter arguments
