---
title: Getting started with fmrigds
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Getting started with fmrigds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("albersdown", quietly = TRUE)) {
  if (requireNamespace("systemfonts", quietly = TRUE)) try(albersdown::albers_register_fonts(), silent = TRUE)
  if (requireNamespace("ggplot2", quietly = TRUE)) try(ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset)), silent = TRUE)
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 6,
  fig.height = 4
)
library(fmrigds)
set.seed(42)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

## Why fmrigds?

Group-level fMRI analysis follows a predictable pattern: you load one or more
subject-level summaries---effect estimates, standardized effects, t-statistics,
z-scores, or p-values---run a group model or combine evidence, and correct for
multiple comparisons. The statistical operations are well-understood, but the
code ends up different every time because the available summaries and input
format vary---CSV, NIfTI, HDF5---and the bookkeeping around subjects,
contrasts, and spatial locations is tedious.

fmrigds gives you a single pipeline interface that works regardless of input
format. You describe *what* you want (fit an unweighted group model, pool
effects with known uncertainty, combine evidence, apply FDR correction) and
fmrigds handles the rest.

## What is a GDS?

A GDS (Group Data Set) is fmrigds's core data structure. It organises
first-level fMRI outputs along three axes:

- **Samples** (rows): the spatial units you measured---voxels, parcels, ROIs,
  or latent components.
- **Subjects** (columns): the participants in your study.
- **Contrasts** (slices): the experimental conditions or comparisons
  (e.g., "faces > places").

Each combination of these three axes can hold one or more **assays**---named
numeric arrays that store the actual data. Common assays include `beta`
(effect estimates), `var` (variance), `t` (t-statistics), `z` (z-scores),
and `p` (p-values). You can think of a GDS as a structured container that
keeps all of these arrays aligned along the same axes.

A GDS also carries a **space** descriptor that records what the sample axis
represents (voxel coordinates, parcel labels, etc.), so spatial information
is never lost.

### Which path fits the summaries you have?

You do not need a variance or standard error to start. Choose the reducer by
what each subject contributes:

| Subject-level input | Ingest as | Group path | What the result means |
|---|---|---|---|
| Effect estimate plus `var` or `se` | `beta` plus `var`/`se` | `"fixed"`, `"random"`, or meta-regression | A pooled effect with first-level precision |
| Beta, standardized effect, or another commensurate signed metric without uncertainty | `beta` | `one_sample()`, `group_ols()`, or `"perm:onesample"` | An unweighted model of the subject-level values |
| z-score | `z` | `"stouffer"` | Combined signed evidence; no pooled effect size |
| t-statistic plus degrees of freedom | `t` plus `df`, then `derive("z")` | `"stouffer"` | Combined signed evidence after calibrating each t-statistic |
| p-value | `p` | `"fisher"` or `"combine:lancaster"` | Combined unsigned evidence; no effect direction |

These paths answer different questions. In particular, a z-score or p-value
does not contain enough information to reconstruct an effect estimate and its
sampling variance. Do not invent unit variance and feed it to a
variance-weighted reducer.

## Plan first, compute later

fmrigds uses **lazy evaluation**: when you call functions like `gds()`,
`derive()`, or `reduce()`, nothing is computed immediately. Instead, you
build up a *plan*---a description of the analysis you want to run. You can
inspect the plan, modify it, or add more steps. When you are satisfied,
calling `compute()` executes the entire pipeline in one pass.

This design has two advantages:

1. You can review the full analysis before any data is loaded.
2. fmrigds can optimise the execution (e.g., only reading the assays it
   needs from disk).

## Quick example

Let's walk through a minimal analysis. First, we create a small CSV file
with ROI-level effect estimates (`beta`) and variances (`var`) for three
subjects:

```{r make-csv}
tmp <- tempfile(fileext = ".csv")
writeLines(c(
  "sample,subject,contrast,beta,var",
  "ROI_1,sub-01,faces,0.5,0.04",
  "ROI_2,sub-01,faces,0.8,0.09",
  "ROI_1,sub-02,faces,1.0,0.04",
  "ROI_2,sub-02,faces,1.2,0.09",
  "ROI_1,sub-03,faces,0.7,0.04",
  "ROI_2,sub-03,faces,0.9,0.09"
), tmp)
```

Each row records the effect estimate and variance for one ROI, one subject,
and one contrast. In a real study, this file might have thousands of rows
covering many ROIs and contrasts.

Now we can run a group-level fixed-effects analysis in three steps:

```{r quick-example}
result <- gds(tmp) |>       # 1. point to the data (creates a plan)
  reduce(method = "fixed") |> # 2. add a group-level reduction step
  compute()                   # 3. execute the plan

assay(result, "beta")
```

```{r quick-example-plot, echo = FALSE, fig.cap = "Group-level effect estimates pooled across three subjects.", fig.height = 3}
b <- drop(assay(result, "beta"))
se <- sqrt(drop(assay(result, "var")))
nms <- rownames(assay(result, "beta"))
if (is.null(nms)) nms <- paste0("ROI_", seq_along(b))

bp <- barplot(b, names.arg = nms, ylim = c(0, max(b + se) * 1.2),
        col = "#5B8DBE", border = NA,
        ylab = "Group beta", main = "Fixed-effects group estimates")
arrows(bp, b - se, bp, b + se, angle = 90, code = 3, length = 0.08, lwd = 1.5)
```

The `|>` operator (called the "pipe") passes the output of one step as the
input to the next, so you can read the pipeline left to right: load the
data, reduce across subjects, then compute.

The result is a group-level GDS containing a single "meta" subject. The
`assay()` function extracts a specific assay by name. Here, `beta` holds
the group-level effect estimates: one value per ROI, pooled across all three
subjects. The error bars show the standard error of the pooled estimate.

## Building and inspecting a plan

Let's look at the plan step by step. Calling `gds()` with a file path
creates a plan that knows *where* the data is, but has not loaded it yet:

```{r build-plan}
plan <- gds(tmp)
explain(plan)
```

The `explain()` function prints a human-readable summary of any plan or
realised GDS. At this stage, the plan only has a source node.

You can chain additional verbs to describe your analysis. Each verb adds
a step to the plan without running anything:

```{r chain-verbs}
plan <- gds(tmp) |>
  derive("t") |>              # compute t-statistics from beta and var
  reduce(method = "fixed") |> # pool across subjects
  posthoc("fdr:bh")           # apply FDR correction

explain(plan)
```

The plan now has four steps. When you are ready, `compute()` runs them all:

```{r compute-plan}
g <- compute(plan)
explain(g)
```

After `compute()`, you have a **realised GDS**---a concrete object holding
arrays of numbers rather than a plan of future operations.

## Exploring a realised GDS

A realised GDS has several accessor functions for inspecting its contents.
Let's start with a simple GDS that has not been reduced yet, so we can see
the full three-dimensional structure:

```{r explore-gds}
g0 <- compute(gds(tmp))
```

**Assays** are the named numeric arrays inside a GDS. You can list them and
extract individual ones:

```{r explore-assays}
# What assays are available?
names(assays(g0))

# Extract the beta array
assay(g0, "beta")
```

The beta array has dimensions **samples x subjects x contrasts**: 2 ROIs,
3 subjects, 1 contrast.

```{r explore-dims}
dim(assay(g0, "beta"))
```

```{r explore-heatmap, echo = FALSE, fig.cap = "Effect estimates (beta) for each ROI and subject.", fig.height = 3}
mat <- assay(g0, "beta")[, , 1]
par(mar = c(4, 5, 3, 1))
image(seq_len(ncol(mat)), seq_len(nrow(mat)), t(mat),
      col = hcl.colors(20, "YlOrRd", rev = TRUE),
      xlab = "Subject", ylab = "", axes = FALSE,
      main = "Beta values (faces contrast)")
axis(1, at = seq_len(ncol(mat)), labels = subjects(g0), las = 1, cex.axis = 0.9)
axis(2, at = seq_len(nrow(mat)),
     labels = rownames(mat) %||% paste0("ROI_", seq_len(nrow(mat))),
     las = 1, cex.axis = 0.9)
# Add numeric values to cells
for (i in seq_len(ncol(mat))) {
  for (j in seq_len(nrow(mat))) {
    text(i, j, sprintf("%.2f", mat[j, i]), cex = 0.9)
  }
}
```

Each cell shows the effect estimate for one ROI in one subject. In a real
study with hundreds of ROIs, you would see spatial patterns emerge across
the heatmap.

You can also query the axis labels directly:

```{r explore-axes}
subjects(g0)
contrasts(g0)
```

The **space** descriptor tells you what the sample axis represents:

```{r explore-space}
space(g0)
```

In this case, the samples are labelled ROIs (ROI_1 and ROI_2). For
voxelwise data, the space would instead describe a 3D grid with an affine
matrix.

### Tidy export

If you want a long-form data frame (useful for plotting with ggplot2 or
exporting to other tools), use `gds_to_tibble()`:

```{r tidy-export}
head(gds_to_tibble(g0))
```

Each row is one observation: one sample, one subject, one contrast, with
columns for each assay.

## Deriving statistics

First-level outputs often include only some statistics. For example, you
might have betas and variances but not t-statistics. The `derive()` verb
fills in the gaps automatically.

```{r derive-stats}
g_derived <- gds(tmp) |>
  derive("t") |>
  compute()

names(assays(g_derived))
```

Asking for `"t"` produced the t-statistic from the existing `beta` and
`var` assays using the formula `t = beta / sqrt(var)`.

```{r derive-plot, echo = FALSE, fig.cap = "Relationship between effect estimates and derived t-statistics.", fig.height = 3.5}
b_vec <- as.vector(assay(g_derived, "beta"))
t_vec <- as.vector(assay(g_derived, "t"))
par(mar = c(4, 4, 3, 1))
plot(b_vec, t_vec, pch = 19, col = "#5B8DBE", cex = 1.3,
     xlab = "Beta (effect estimate)", ylab = "t-statistic",
     main = "Derived t from beta and var")
abline(lm(t_vec ~ b_vec), lty = 2, col = "grey40")
```

ROIs with larger effects *and* smaller variances produce higher
t-statistics. The `derive()` verb resolves dependencies
automatically---if you ask for `"z"` but only have `beta` and `var`, it
will first derive `t`, then convert to `z`.

Here are the available derivations:

| Target | Requirements    | What it computes                 |
|--------|-----------------|----------------------------------|
| `t`    | beta, var       | `beta / sqrt(var)`               |
| `z`    | t (or beta+var) | z-score (normal approximation)   |
| `p`    | z (or t)        | two-sided p-value                |
| `se`   | var             | `sqrt(var)` (standard error)     |
| `var`  | se              | `se^2` (variance)                |

Derivation only follows mathematically identified relationships. It cannot
recover a missing effect estimate or sampling variance from a z-score,
t-statistic, or p-value alone.

## Group-level analysis with reduce()

The `reduce()` verb is where the group-level analysis happens. It combines
data across subjects into a single summary, much like a meta-analysis
combines results across studies.

### Effects without first-level variance

If every subject contributes the same kind of signed, commensurate quantity
but no standard error---a beta, standardized effect, AUC contrast, or another
scalar map---fit an unweighted model. Map the value column to the GDS `beta`
assay, which is the response slot used by the group-model reducers:

```{r reduce-effect-only}
effect_only <- data.frame(
  sample = rep(c("ROI_1", "ROI_2"), 3),
  subject = rep(c("sub-01", "sub-02", "sub-03"), each = 2),
  contrast = "faces",
  effect_size = c(0.5, 0.8, 1.0, 1.2, 0.7, 0.9)
)

g_effect <- gds(effect_only, effect_cols = list(beta = "effect_size")) |>
  one_sample() |>
  compute()

assay(g_effect, "coef:(Intercept)")
```

Here, uncertainty is estimated from between-subject variation. Use
`group_ols()` instead when you need subject covariates, or
`reduce(method = "perm:onesample")` for the sign-flip permutation path. If
your effect size has a valid sampling standard error, ingest it as `se`, derive
`var`, and use a meta-analytic reducer instead, for example
`gds(data, effect_cols = list(beta = "effect_size", se = "se")) |>
derive("var") |> reduce("random")`.

### z, t, or p summary statistics

When you have test statistics rather than effects, combine evidence directly.
For z-scores, name the input assay explicitly and use Stouffer's method:

```{r reduce-z-only}
z_only <- transform(
  effect_only[c("sample", "subject", "contrast")],
  z = c(1.5, 2.0, 2.1, 2.4, 1.8, 2.2)
)

g_z <- gds(z_only, effect_cols = list(z = "z")) |>
  reduce(method = "stouffer", weights = "equal") |>
  compute()

assay(g_z, "z")
assay(g_z, "p")
```

A t-statistic needs its degrees of freedom before it can be calibrated as a
z-score. Once both columns are present, the rest of the path is the same:

```{r reduce-t-only}
t_only <- transform(
  z_only[c("sample", "subject", "contrast")],
  t = z_only$z,
  df = 30
)

g_t <- gds(t_only, effect_cols = list(t = "t", df = "df")) |>
  derive("z") |>
  reduce(method = "stouffer", weights = "equal") |>
  compute()

assay(g_t, "p")
```

For p-values, use `effect_cols = list(p = "p")` with `method = "fisher"`
or `method = "combine:lancaster"`. Fisher's method does not retain effect
direction; prefer Stouffer's method when signed evidence matters.

### Fixed-effects

A fixed-effects model assumes that every subject shares the same true
effect. It pools estimates using inverse-variance weighting---subjects with
more precise estimates (smaller variance) get more weight:

```{r reduce-fe}
g_fe <- gds(tmp) |>
  reduce(method = "fixed") |>
  compute()

names(assays(g_fe))
```

The result contains the pooled `beta`, its `var`, derived `z` and `p`
values, and diagnostic assays like `cochran_q` (a test of whether the
effect really is the same across subjects). Notice that `subjects()` now
returns `"meta"` instead of individual subject names:

```{r reduce-fe-subjects}
subjects(g_fe)
```

### Random-effects

If you expect the true effect to vary across subjects (which is common in
fMRI), a random-effects model is more appropriate. It estimates both the
average effect and the between-subject variability:

```{r reduce-re}
g_re <- gds(tmp) |>
  reduce(method = "random") |>
  compute()

names(assays(g_re))
```

The additional `tau2` assay estimates between-subject variance at each
sample location. When `tau2` is large, subjects disagree about the
effect size at that location.

```{r reduce-compare, echo = FALSE, fig.cap = "Fixed- vs random-effects group estimates. Error bars show standard errors.", fig.height = 3.5}
b_fe <- drop(assay(g_fe, "beta"))
b_re <- drop(assay(g_re, "beta"))
se_fe <- sqrt(drop(assay(g_fe, "var")))
se_re <- sqrt(drop(assay(g_re, "var")))
nms <- rownames(assay(g_fe, "beta"))
if (is.null(nms)) nms <- paste0("ROI_", seq_along(b_fe))

x <- seq_along(b_fe)
par(mar = c(4, 4, 3, 1))
plot(x - 0.1, b_fe, pch = 19, col = "#5B8DBE", cex = 1.4,
     xlim = c(0.5, length(x) + 0.5), ylim = range(c(b_fe - se_fe, b_re + se_re)) * c(0.8, 1.2),
     xlab = "", ylab = "Group beta", main = "Fixed vs random effects",
     xaxt = "n")
points(x + 0.1, b_re, pch = 17, col = "#D4694A", cex = 1.4)
arrows(x - 0.1, b_fe - se_fe, x - 0.1, b_fe + se_fe,
       angle = 90, code = 3, length = 0.06, col = "#5B8DBE", lwd = 1.5)
arrows(x + 0.1, b_re - se_re, x + 0.1, b_re + se_re,
       angle = 90, code = 3, length = 0.06, col = "#D4694A", lwd = 1.5)
axis(1, at = x, labels = nms)
legend("topleft", pch = c(19, 17), col = c("#5B8DBE", "#D4694A"),
       legend = c("Fixed", "Random"), bty = "n")
```

Random-effects standard errors are wider because they account for
between-subject variability in addition to within-subject measurement error.

### Meta-regression

You can include subject-level covariates (e.g., age, group membership) in
the group model. First attach the covariates with `with_col_data()`, then
specify a formula:

```{r reduce-reg}
covariates <- data.frame(
  age = c(25, 32, 28),
  row.names = c("sub-01", "sub-02", "sub-03")
)

g_reg <- gds(tmp) |>
  with_col_data(covariates) |>
  reduce(method = "meta:fe_reg", formula = ~ age) |>
  compute()

names(assays(g_reg))
```

Each model coefficient and its standard error becomes a separate assay.
Here, `coef:age` tells you how the effect changes per year of age at each
ROI.

### Repeated-measures mixed models

For repeated-measures designs, `fmrigds` also provides a restricted Gaussian
mixed-model path that is optimized for the case where every ROI or voxel shares
the same observation layout and the same fixed/random design. The repeated
measure lives on the contrast axis, and its metadata can be extracted directly
from a stacked long table:

```{r reduce-lmm}
subjects_rm <- paste0("sub-", sprintf("%02d", 1:4))
visits_rm <- c("pre", "mid", "post")
time_rm <- c(-1, 0, 1)

rm_df <- expand.grid(
  sample = c("ROI_1", "ROI_2"),
  subject = subjects_rm,
  contrast = visits_rm,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)

subject_intercept <- c(-0.20, -0.05, 0.10, 0.25)
names(subject_intercept) <- subjects_rm
sample_shift <- c(ROI_1 = 0.00, ROI_2 = 0.30)
sample_slope <- c(ROI_1 = 0.45, ROI_2 = 0.65)

rm_df$time <- time_rm[match(rm_df$contrast, visits_rm)]
rm_df$beta <- 0.8 +
  sample_shift[rm_df$sample] +
  subject_intercept[rm_df$subject] +
  sample_slope[rm_df$sample] * rm_df$time
rm_df$var <- 0.04

g_rm <- gds(rm_df, contrast_data_cols = "time") |>
  reduce(
    method = "lmm:ri_slope1_knownvar",
    formula = ~ time,
    options = list(
      slope = "time",
      covariance = "diag",
      fit = "REML"
    )
  ) |>
  compute()

names(assays(g_rm))
```

The fixed-effect outputs follow the same `coef:*`, `se_coef:*`, `t_coef:*`,
and `p_coef:*` naming pattern as the regression reducers. Additional assays
report variance components such as `vc_intercept`, `vc_slope`,
`vc_resid`, and `sampling_var_mean`.

Four restricted mixed-model reducers are available:

- `method = "lmm:ri"` for a random intercept only
- `method = "lmm:ri_slope1"` for a random intercept plus one within-subject slope
- `method = "lmm:ri_knownvar"` and `"lmm:ri_slope1_knownvar"` for the same
  random-effects structures with `var` included as known diagonal sampling
  variance

The unweighted methods support pooled or voxelwise variance parameters. The
known-variance methods are currently voxelwise only and fit
`diag(var) + vc_resid * I + Z G Z'`. If only effect estimates are available,
use an unweighted method; do not manufacture unit variances. See
`vignette("repeated-measures-lmm")` for implications and variance-quality
checks. This family remains intentionally narrower than `lmer()`.

## Post-hoc corrections

When you test many spatial locations simultaneously (hundreds of ROIs,
thousands of voxels), some will appear significant by chance. Post-hoc
corrections control the false discovery rate.

Add a correction step with `posthoc()`:

```{r posthoc-fdr}
g_fdr <- gds(tmp) |>
  reduce(method = "fixed") |>
  posthoc("fdr:bh") |>
  compute()

# The original p-values
assay(g_fdr, "p")

# FDR-adjusted q-values
assay(g_fdr, "q")
```

A q-value of 0.05 means that among all locations with q-values at or below
0.05, you expect no more than 5% to be false discoveries. Available
correction methods:

- `"fdr:bh"` --- Benjamini--Hochberg (standard FDR; the default choice)
- `"fdr:by"` --- Benjamini--Yekutieli (valid under arbitrary dependence,
  but more conservative)
- `"fdr:spatial"` --- spatially-aware FDR that groups nearby tests; see
  `vignette("as-plan-and-spatial-fdr")`

## Writing results to disk

The `write_out()` verb saves results as part of the pipeline. It runs
transparently---the data passes through to the next step, and a copy is
written to the path you specify:

```{r write-csv}
out_csv <- tempfile(fileext = ".csv")

gds(tmp) |>
  reduce(method = "fixed") |>
  write_out(out_csv, format = "csv") |>
  compute()

head(read.csv(out_csv))
```

Supported formats:

| Format      | Extension | Best for                          |
|-------------|-----------|-----------------------------------|
| `"csv"`     | .csv      | Sharing, inspection in Excel      |
| `"parquet"` | .parquet  | Fast columnar storage             |
| `"h5"`      | .h5       | Native fmrigds archive (`/gds`)   |
| `"nifti"`   | .nii.gz   | Voxelwise maps for neuroimaging tools |

HDF5 is a container, not a synonym for the legacy fmristore schema.
`write_out(..., format = "h5")` writes the current native fmrigds `/gds`
layout. The writer stores assays, axes, spatial structure, axis tables,
registered map families, metadata, and provenance records. The current reader
restores the supported GDS data model; it does not promise that every arbitrary
custom metadata field will be reconstructed by every fmrigds version.

```{r write-h5, eval = FALSE}
gds(tmp) |>
  reduce(method = "fixed") |>
  write_out("group_results.h5", format = "h5") |>
  compute()

# Later: read the native /gds archive (not a legacy fmristore layout)
plan <- gds("group_results.h5")
restored <- compute(plan)
```

If exact preservation of an application-specific metadata field matters,
verify that field after reading with the fmrigds version used by your workflow.
The legacy label in `vignette("fmristore-ingestion")` applies only to older
fmristore layouts such as LabeledVolumeSet, H5ParcellatedScan, and
LatentNeuroVec.

## Working with NIfTI files

For voxelwise analyses, you typically have one NIfTI file per subject per
contrast, spread across a directory tree. The `image_catalog()` function
discovers and organises these files automatically:

```{r nifti-catalog, eval = FALSE}
# Discover all NIfTI files under a study directory
catalog <- image_catalog(
  root = "study_dir",
  pattern = "sub-*/stats/*.nii.gz",
  path_regex = "sub-(?<subject>[^/]+)"
)

# Tell fmrigds which files are betas and which are standard errors
catalog <- catalog |>
  map_assays(
    beta = "cope[0-9]+\\.nii",
    se   = "varcope[0-9]+\\.nii"
  )

# Convert to a GDS plan, applying a group mask
plan <- as_gds(catalog, mask = "group_mask.nii.gz")

# Run the group analysis
result <- plan |> reduce(method = "fixed") |> compute()
```

The code above cannot run here because it needs real NIfTI files on disk.
See `?image_catalog` for file discovery and `?map_assays` for mapping
filenames to assay types.

## The pipeline at a glance

Here is the beta-plus-variance branch from start to finish:

```
  gds("data.csv")
    |> derive("t")
    |> mask()
    |> reduce(method = "fixed")
    |> posthoc("fdr:bh")
    |> write_out("results.csv", format = "csv")
    |> compute()
```

For effect-only inputs, replace `derive("t") |> reduce(method = "fixed")`
with `one_sample()`. For z-scores, replace it with
`reduce(method = "stouffer", weights = "equal")`. Each verb adds one node to
the plan, and `compute()` executes them in sequence. With `r 2` ROIs this runs
instantly; in a typical whole-brain analysis with 100,000+ voxels, the same
pipeline scales because fmrigds reads data lazily and streams through each
step.

## Next steps

Now that you understand the core pipeline, you might want to explore:

- `vignette("spatial-operations")` --- masking missing data, aligning
  subjects in latent spaces, and transforming between spatial
  representations
- `vignette("as-plan-and-spatial-fdr")` --- chaining operations on realised
  GDS objects and spatially-aware FDR correction
- `vignette("fmristore-ingestion")` --- reading native fmrigds HDF5 archives
  and older fmristore layouts (labeled volumes, parcellated data, latent bases)
- `?reduce` for the full list of reduction methods, including OLS, Stouffer,
  Fisher, and Lancaster

```{r cleanup, include = FALSE}
unlink(c(tmp, out_csv))
```
