---
title: "Getting started with fmrigds"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with fmrigds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(fmrigds)
set.seed(42)
```

## Why fmrigds?

Group-level fMRI analysis follows a predictable pattern: you load first-level
results (effect estimates, variances) for each subject, compute any missing
statistics, run a group model, and correct for multiple comparisons. The
statistical operations are well-understood, but the code ends up different
every time because the input format varies---CSV, NIfTI, HDF5---and the
bookkeeping around subjects, contrasts, and spatial locations is tedious.

fmrigds gives you a single pipeline interface that works regardless of input
format. You describe *what* you want (derive t-statistics, run a fixed-effects
model, apply FDR correction) and fmrigds handles the rest.

## What is a GDS?

A GDS (Group Data Set) is fmrigds's core data structure. It organises
first-level fMRI outputs along three axes:

- **Samples** (rows): the spatial units you measured---voxels, parcels, ROIs,
  or latent components.
- **Subjects** (columns): the participants in your study.
- **Contrasts** (slices): the experimental conditions or comparisons
  (e.g., "faces > places").

Each combination of these three axes can hold one or more **assays**---named
numeric arrays that store the actual data. Common assays include `beta`
(effect estimates), `var` (variance), `t` (t-statistics), `z` (z-scores),
and `p` (p-values). You can think of a GDS as a structured container that
keeps all of these arrays aligned along the same axes.

A GDS also carries a **space** descriptor that records what the sample axis
represents (voxel coordinates, parcel labels, etc.), so spatial information
is never lost.

## Plan first, compute later

fmrigds uses **lazy evaluation**: when you call functions like `gds()`,
`derive()`, or `reduce()`, nothing is computed immediately. Instead, you
build up a *plan*---a description of the analysis you want to run. You can
inspect the plan, modify it, or add more steps. When you are satisfied,
calling `compute()` executes the entire pipeline in one pass.

This design has two advantages:

1. You can review the full analysis before any data is loaded.
2. fmrigds can optimise the execution (e.g., only reading the assays it
   needs from disk).

## Quick example

Let's walk through a minimal analysis. First, we create a small CSV file
with ROI-level effect estimates (`beta`) and variances (`var`) for three
subjects:

```{r make-csv}
tmp <- tempfile(fileext = ".csv")
writeLines(c(
  "sample,subject,contrast,beta,var",
  "ROI_1,sub-01,faces,0.5,0.04",
  "ROI_2,sub-01,faces,0.8,0.09",
  "ROI_1,sub-02,faces,1.0,0.04",
  "ROI_2,sub-02,faces,1.2,0.09",
  "ROI_1,sub-03,faces,0.7,0.04",
  "ROI_2,sub-03,faces,0.9,0.09"
), tmp)
```

Each row records the effect estimate and variance for one ROI, one subject,
and one contrast. In a real study, this file might have thousands of rows
covering many ROIs and contrasts.

Now we can run a group-level fixed-effects analysis in three steps:

```{r quick-example}
result <- gds(tmp) |>       # 1. point to the data (creates a plan)
  reduce(method = "fixed") |> # 2. add a group-level reduction step
  compute()                   # 3. execute the plan

assay(result, "beta")
```

The `|>` operator (called the "pipe") passes the output of one step as the
input to the next, so you can read the pipeline left to right: load the
data, reduce across subjects, then compute.

The result is a group-level GDS containing a single "meta" subject. The
`assay()` function extracts a specific assay by name. Here, `beta` holds
the group-level effect estimates: one value per ROI, pooled across all three
subjects.

## Building and inspecting a plan

Let's look at the plan step by step. Calling `gds()` with a file path
creates a plan that knows *where* the data is, but has not loaded it yet:

```{r build-plan}
plan <- gds(tmp)
explain(plan)
```

The `explain()` function prints a human-readable summary of any plan or
realised GDS. At this stage, the plan only has a source node.

You can chain additional verbs to describe your analysis. Each verb adds
a step to the plan without running anything:

```{r chain-verbs}
plan <- gds(tmp) |>
  derive("t") |>              # compute t-statistics from beta and var
  reduce(method = "fixed") |> # pool across subjects
  posthoc("fdr:bh")           # apply FDR correction

explain(plan)
```

The plan now has four steps. When you are ready, `compute()` runs them all:

```{r compute-plan}
g <- compute(plan)
explain(g)
```

After `compute()`, you have a **realised GDS**---a concrete object holding
arrays of numbers rather than a plan of future operations.

## Exploring a realised GDS

A realised GDS has several accessor functions for inspecting its contents.
Let's start with a simple GDS that has not been reduced yet, so we can see
the full three-dimensional structure:

```{r explore-gds}
g0 <- compute(gds(tmp))
```

**Assays** are the named numeric arrays inside a GDS. You can list them and
extract individual ones:

```{r explore-assays}
# What assays are available?
names(assays(g0))

# Extract the beta array
assay(g0, "beta")
```

The beta array has dimensions **samples x subjects x contrasts**: 2 ROIs,
3 subjects, 1 contrast.

```{r explore-dims}
dim(assay(g0, "beta"))
```

You can also query the axis labels directly:

```{r explore-axes}
subjects(g0)
contrasts(g0)
```

The **space** descriptor tells you what the sample axis represents:

```{r explore-space}
space(g0)
```

In this case, the samples are labelled ROIs (ROI_1 and ROI_2). For
voxelwise data, the space would instead describe a 3D grid with an affine
matrix.

### Tidy export

If you want a long-form data frame (useful for plotting with ggplot2 or
exporting to other tools), use `gds_to_tibble()`:

```{r tidy-export}
head(gds_to_tibble(g0))
```

Each row is one observation: one sample, one subject, one contrast, with
columns for each assay.

## Deriving statistics

First-level outputs often include only some statistics. For example, you
might have betas and variances but not t-statistics. The `derive()` verb
fills in the gaps automatically.

```{r derive-stats}
g_derived <- gds(tmp) |>
  derive("t") |>
  compute()

names(assays(g_derived))
```

Asking for `"t"` produced the t-statistic from the existing `beta` and
`var` assays using the formula `t = beta / sqrt(var)`. fmrigds knows the
standard relationships between statistics and resolves dependencies
automatically---if you ask for `"z"` but only have `beta` and `var`, it
will first derive `t`, then convert to `z`.

Here are the available derivations:

| Target | Requirements    | What it computes                 |
|--------|-----------------|----------------------------------|
| `t`    | beta, var       | `beta / sqrt(var)`               |
| `z`    | t (or beta+var) | z-score (normal approximation)   |
| `p`    | z (or t)        | two-sided p-value                |
| `se`   | var             | `sqrt(var)` (standard error)     |
| `var`  | se              | `se^2` (variance)                |

## Group-level analysis with reduce()

The `reduce()` verb is where the group-level analysis happens. It combines
data across subjects into a single summary, much like a meta-analysis
combines results across studies.

### Fixed-effects

A fixed-effects model assumes that every subject shares the same true
effect. It pools estimates using inverse-variance weighting---subjects with
more precise estimates (smaller variance) get more weight:

```{r reduce-fe}
g_fe <- gds(tmp) |>
  reduce(method = "fixed") |>
  compute()

names(assays(g_fe))
```

The result contains the pooled `beta`, its `var`, derived `z` and `p`
values, and diagnostic assays like `cochran_q` (a test of whether the
effect really is the same across subjects). Notice that `subjects()` now
returns `"meta"` instead of individual subject names:

```{r reduce-fe-subjects}
subjects(g_fe)
```

### Random-effects

If you expect the true effect to vary across subjects (which is common in
fMRI), a random-effects model is more appropriate. It estimates both the
average effect and the between-subject variability:

```{r reduce-re}
g_re <- gds(tmp) |>
  reduce(method = "random") |>
  compute()

names(assays(g_re))
```

The additional `tau2` assay estimates between-subject variance at each
sample location. When `tau2` is large, subjects disagree about the
effect size at that location.

### Meta-regression

You can include subject-level covariates (e.g., age, group membership) in
the group model. First attach the covariates with `with_col_data()`, then
specify a formula:

```{r reduce-reg}
covariates <- data.frame(
  age = c(25, 32, 28),
  row.names = c("sub-01", "sub-02", "sub-03")
)

g_reg <- gds(tmp) |>
  with_col_data(covariates) |>
  reduce(method = "meta:fe_reg", formula = ~ age) |>
  compute()

names(assays(g_reg))
```

Each model coefficient and its standard error becomes a separate assay.
Here, `coef:age` tells you how the effect changes per year of age at each
ROI.

## Post-hoc corrections

When you test many spatial locations simultaneously (hundreds of ROIs,
thousands of voxels), some will appear significant by chance. Post-hoc
corrections control the false discovery rate.

Add a correction step with `posthoc()`:

```{r posthoc-fdr}
g_fdr <- gds(tmp) |>
  reduce(method = "fixed") |>
  posthoc("fdr:bh") |>
  compute()

# The original p-values
assay(g_fdr, "p")

# FDR-adjusted q-values
assay(g_fdr, "q")
```

A q-value of 0.05 means that among all locations with q-values at or below
0.05, you expect no more than 5% to be false discoveries. Available
correction methods:

- `"fdr:bh"` --- Benjamini--Hochberg (standard FDR; the default choice)
- `"fdr:by"` --- Benjamini--Yekutieli (valid under arbitrary dependence,
  but more conservative)
- `"fdr:spatial"` --- spatially-aware FDR that groups nearby tests; see
  `vignette("as-plan-and-spatial-fdr")`

## Writing results to disk

The `write_out()` verb saves results as part of the pipeline. It runs
transparently---the data passes through to the next step, and a copy is
written to the path you specify:

```{r write-csv}
out_csv <- tempfile(fileext = ".csv")

gds(tmp) |>
  reduce(method = "fixed") |>
  write_out(out_csv, format = "csv") |>
  compute()

head(read.csv(out_csv))
```

Supported formats:

| Format      | Extension | Best for                          |
|-------------|-----------|-----------------------------------|
| `"csv"`     | .csv      | Sharing, inspection in Excel      |
| `"parquet"` | .parquet  | Fast columnar storage             |
| `"h5"`      | .h5       | Archiving with full metadata      |
| `"nifti"`   | .nii.gz   | Voxelwise maps for neuroimaging tools |

For archiving with full provenance, use HDF5. These files round-trip
perfectly---reading them back with `gds()` recovers all metadata:

```{r write-h5, eval = FALSE}
gds(tmp) |>
  reduce(method = "fixed") |>
  write_out("group_results.h5", format = "h5") |>
  compute()

# Later: read back with full fidelity
plan <- gds("group_results.h5")
```

## Working with NIfTI files

For voxelwise analyses, you typically have one NIfTI file per subject per
contrast, spread across a directory tree. The `image_catalog()` function
discovers and organises these files automatically:

```{r nifti-catalog, eval = FALSE}
# Discover all NIfTI files under a study directory
catalog <- image_catalog(
  root = "study_dir",
  pattern = "sub-*/stats/*.nii.gz",
  path_regex = "sub-(?<subject>[^/]+)"
)

# Tell fmrigds which files are betas and which are standard errors
catalog <- catalog |>
  map_assays(
    beta = "cope[0-9]+\\.nii",
    se   = "varcope[0-9]+\\.nii"
  )

# Convert to a GDS plan, applying a group mask
plan <- as_gds(catalog, mask = "group_mask.nii.gz")

# Run the group analysis
result <- plan |> reduce(method = "fixed") |> compute()
```

The code above cannot run here because it needs real NIfTI files on disk.
See `?image_catalog` for file discovery and `?map_assays` for mapping
filenames to assay types.

## Next steps

Now that you understand the core pipeline, you might want to explore:

- `vignette("spatial-operations")` --- masking missing data, aligning
  subjects in latent spaces, and transforming between spatial
  representations
- `vignette("as-plan-and-spatial-fdr")` --- chaining operations on realised
  GDS objects and spatially-aware FDR correction
- `vignette("fmristore-ingestion")` --- reading fmristore HDF5 files
  (labeled volumes, parcellated data, latent bases)
- `?reduce` for the full list of reduction methods, including OLS, Stouffer,
  Fisher, and Lancaster

```{r cleanup, include = FALSE}
unlink(c(tmp, out_csv))
```
