---
title: "Getting started with fmrigds"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with fmrigds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(fmrigds)
set.seed(42)
```

## Why fmrigds?

Group-level fMRI analysis follows a predictable pattern: load first-level
results (betas, variances) for each subject, derive any missing statistics,
run a group model, and correct for multiple comparisons. The statistical
operations are well-understood, but the code ends up different every time
because the input format varies---CSV, NIfTI, HDF5---and the bookkeeping is
tedious.

fmrigds gives you a single pipeline interface that works regardless of input
format. You describe *what* you want (derive t-statistics, run a fixed-effects
model, apply FDR correction) and fmrigds handles the rest.

The design centres on **lazy evaluation**: you build a plan describing your
analysis, then execute it with `compute()`. This lets you inspect and modify the
pipeline before any data is loaded.

## Quick example

```{r quick-example}
# A small CSV with ROI-level betas and variances for 3 subjects
tmp <- tempfile(fileext = ".csv")
writeLines(c(
  "sample,subject,contrast,beta,var",
  "ROI_1,sub-01,faces,0.5,0.04",
  "ROI_2,sub-01,faces,0.8,0.09",
  "ROI_1,sub-02,faces,1.0,0.04",
  "ROI_2,sub-02,faces,1.2,0.09",
  "ROI_1,sub-03,faces,0.7,0.04",
  "ROI_2,sub-03,faces,0.9,0.09"
), tmp)

# Build a plan, reduce across subjects, compute
result <- gds(tmp) |>
  reduce(method = "fixed") |>
  compute()

# Group-level effect estimates
assay(result, "beta")
```

Three lines of analysis turn a CSV into a group-level GDS with effect
estimates, variances, z-scores, p-values, and heterogeneity diagnostics.

## Building a plan

Point `gds()` at your data file. The format is auto-detected:

```{r build-plan}
plan <- gds(tmp)
explain(plan)
```

Nothing is loaded into memory yet. Chain verbs to describe your pipeline:

```{r chain-verbs}
plan <- gds(tmp) |>
  derive("t") |>
  reduce(method = "fixed") |>
  posthoc("fdr:bh")

explain(plan)
```

When you're ready, `compute()` executes the full pipeline:

```{r compute-plan}
g <- compute(plan)
explain(g)
```

## Exploring a GDS object

After `compute()`, you have a realized GDS. Use the accessor functions to
inspect it:

```{r explore-gds}
g0 <- compute(gds(tmp))

# What assays are available?
names(assays(g0))

# The beta array: samples x subjects x contrasts
dim(assay(g0, "beta"))

# Subject and contrast identifiers
subjects(g0)
contrasts(g0)

# Spatial layout
space(g0)
```

For a tidy long-form data frame (useful for plotting or export), use
`gds_to_tibble()`:

```{r tidy-export}
head(gds_to_tibble(g0))
```

## Deriving statistics

`derive()` computes new assays from existing ones. fmrigds knows the standard
relationships:

```{r derive-stats}
g_derived <- gds(tmp) |>
  derive("t") |>
  compute()

names(assays(g_derived))
```

Dependencies are resolved automatically. Available derivations:

| Target | Requirements    | Description    |
|--------|-----------------|----------------|
| `t`    | beta, var       | t-statistic    |
| `z`    | t (or beta+var) | z-score        |
| `p`    | z (or t)        | p-value        |
| `se`   | var             | Standard error |
| `var`  | se              | Variance       |

## Group-level analysis

The `reduce()` verb combines data across subjects into group-level summaries.

### Fixed-effects meta-analysis

Assumes a common true effect, weighted by precision (1/variance):

```{r reduce-fe}
g_fe <- gds(tmp) |>
  reduce(method = "fixed") |>
  compute()

names(assays(g_fe))
subjects(g_fe)
```

### Random-effects meta-analysis

Accounts for between-subject heterogeneity using the DerSimonian--Laird
estimator:

```{r reduce-re}
g_re <- gds(tmp) |>
  reduce(method = "random") |>
  compute()

names(assays(g_re))
```

The additional `tau2` assay estimates between-subject variance at each sample
location.

### Meta-regression

Include subject-level covariates with `with_col_data()` and a model formula:

```{r reduce-reg}
covariates <- data.frame(
  age = c(25, 32, 28),
  row.names = c("sub-01", "sub-02", "sub-03")
)

g_reg <- gds(tmp) |>
  with_col_data(covariates) |>
  reduce(method = "meta:fe_reg", formula = ~ age) |>
  compute()

names(assays(g_reg))
```

Each model coefficient and its standard error becomes a separate assay.

## Post-hoc corrections

Apply multiple comparison corrections after group analysis:

```{r posthoc-fdr}
g_fdr <- gds(tmp) |>
  reduce(method = "fixed") |>
  posthoc("fdr:bh") |>
  compute()

# FDR-adjusted q-values
assay(g_fdr, "q")
```

Available methods: `"fdr:bh"` (Benjamini--Hochberg), `"fdr:by"`
(Benjamini--Yekutieli), and `"fdr:spatial"` for spatially-aware correction.
See `vignette("as-plan-and-spatial-fdr")` for the spatial approach.

## Writing results

`write_out()` saves results to disk as part of the pipeline:

```{r write-csv}
out_csv <- tempfile(fileext = ".csv")

gds(tmp) |>
  reduce(method = "fixed") |>
  write_out(out_csv, format = "csv") |>
  compute()

head(read.csv(out_csv))
```

Supported formats: CSV (`"csv"`), HDF5 (`"h5"`), Parquet (`"parquet"`), and
NIfTI (`"nifti"`). For archiving with full metadata and provenance, use HDF5:

```{r write-h5, eval = FALSE}
gds(tmp) |>
  reduce(method = "fixed") |>
  write_out("group_results.h5", format = "h5") |>
  compute()
```

## Working with NIfTI files

For voxelwise analyses, `image_catalog()` discovers and organizes NIfTI files
across subjects:

```{r nifti-catalog, eval = FALSE}
catalog <- image_catalog(
  root = "study_dir",
  pattern = "sub-*/stats/*.nii.gz",
  path_regex = "sub-(?<subject>[^/]+)"
)

catalog <- catalog |>
  map_assays(
    beta = "cope[0-9]+\\.nii",
    se   = "varcope[0-9]+\\.nii"
  )

plan <- as_gds(catalog, mask = "group_mask.nii.gz")
result <- plan |> reduce(method = "fixed") |> compute()
```

See `?image_catalog` for file discovery and `?map_assays` for mapping files to
assay types.

## Next steps

- `vignette("as-plan-and-spatial-fdr")` --- chaining operations on realized GDS
  objects and spatially-aware FDR correction
- `vignette("fmristore-ingestion")` --- reading fmristore HDF5 files (labeled
  volumes, parcellated data, latent bases)
- `?reduce` for the full list of reduction methods including OLS, Stouffer,
  Fisher, and Lancaster

```{r cleanup, include = FALSE}
unlink(c(tmp, out_csv))
```
