---
title: Examine subjects under a group model
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: interaction
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Examine subjects under a group model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("albersdown", quietly = TRUE)) {
  if (requireNamespace("systemfonts", quietly = TRUE)) try(albersdown::albers_register_fonts(), silent = TRUE)
  if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
library(fmrigds)
set.seed(20260805)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-interaction","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```


You have fitted a group model, but its average alone cannot tell you whether a
participant is unexpected for their age, group, or site, or whether that
participant materially changes the scientific result. These are different
questions. A noisy estimate can look unusual and carry little weight; a precise,
otherwise conventional estimate can have substantial influence.

`examine_group()` evaluates those questions against the model already present
in an fmrigds plan. It returns subject summaries, group diagnostic maps, a
model-adjusted residual geometry, retained-subject maps, sensitivity results,
and explicit availability states. It never changes the source analysis.

## What question does the examination answer?

The examination asks what happens when each subject is treated as a prediction
target rather than as evidence for their own expected value. For subject
\(i\), the fitted group model supplies a leave-one-subject-out expectation
\(\hat y_{-i}\). fmrigds then compares the observed subject map with that
expectation and compares the full group result with the result after deleting
that subject.

This produces two deliberately separate questions:

1. **Model surprise:** Is this subject's spatial pattern difficult for the
   other subjects and the specified covariates to predict?
2. **Group influence:** Does deleting this subject materially change the
   requested group statistic?

They should not be collapsed into one generic outlier score. A subject with a
large first-level variance can be surprising but receive little weight. A
precise subject can look ordinary yet materially affect a group coefficient.
A quality concern such as high motion is a third signal: it concerns data
validity, not statistical unusualness.

### What do the main terms mean?

| Term | Quantity summarized over samples | Interpretation |
|---|---|---|
| Predictive residual | Observed effect minus its leave-one-out expectation, scaled by predictive uncertainty | Signed local mismatch after adjustment for the group model |
| Surprise energy | Root mean capped squared predictive residual | Overall magnitude of the model mismatch for one subject and contrast |
| Tail extent | Fraction of eligible samples with an absolute predictive residual above the configured residual threshold | Whether surprise is spatially broad rather than a single extreme value |
| Estimand | A named linear combination of group-model coefficients | The scientific quantity whose sensitivity you want to assess |
| Deletion statistic | Full-data group statistic minus the statistic after deleting one subject | Signed local change in the requested estimand's test statistic |
| Influence energy | Root mean capped squared deletion statistic | Overall sensitivity of that group statistic to one subject |
| Stability | Fraction of deterministic feature splits that reproduce a gate | Protection against a flag driven by one idiosyncratic feature subset |
| Quality criterion | An explicitly directed threshold on subject metadata | A data-validity review signal, kept separate from surprise and influence |

The cap bounds the contribution of any one sample before energies are
averaged. It makes the summaries describe a pattern across the map rather than
allowing one enormous residual to dominate. Tail extent remains available in
the contrast table, the influence table retains the uncapped maximum deletion
statistic, and retained cases keep their spatial residual maps.

## What data do you need?

The most informative examination starts from subject effect estimates and
genuine first-level variances. Subject covariates and first-level quality
metrics belong in `col_data`. The example below has two task contrasts, a group
effect, an age term, two acquisition sites, one unusually precise subject, and
one subject whose first contrast is reversed.

```{r make-example, include=FALSE}
n_subject <- 16L
n_sample <- 72L
subject_id <- sprintf("sub-%03d", seq_len(n_subject))
group <- factor(rep(c("control", "patient"), each = n_subject / 2))
site <- factor(rep(c("site-a", "site-b"), length.out = n_subject))
age <- seq(21, 62, length.out = n_subject)
mean_fd <- c(runif(n_subject - 1L, 0.08, 0.22), 0.56)
participants <- data.frame(group, age, site, mean_fd, row.names = subject_id)

feature <- seq(0, 2 * pi, length.out = n_sample)
patterns <- cbind(sin(feature), cos(feature))
beta <- array(NA_real_, c(n_sample, n_subject, 2L))
for (i in seq_len(n_subject)) {
  group_shift <- if (group[i] == "patient") 0.35 else 0
  age_shift <- 0.006 * (age[i] - mean(age))
  site_shift <- if (site[i] == "site-b") 0.12 else 0
  beta[, i, ] <- patterns + group_shift + age_shift + site_shift +
    matrix(rnorm(n_sample * 2L, sd = 0.12), n_sample, 2L)
}
beta[, 15L, 1L] <- -patterns[, 1L] + rnorm(n_sample, sd = 0.05)
variance <- array(0.08, dim(beta))
variance[, 8L, ] <- 0.006

subject_maps <- new_gds(
  assays = list(beta = beta, var = variance),
  space = space_sample_labels(sprintf("feature-%03d", seq_len(n_sample))),
  subjects = subject_id,
  contrasts = c("faces>places", "tools>faces"),
  col_data = participants
)
```

```{r inspect-input}
subject_maps
head(col_data(subject_maps))
```

## How do you run the examination?

Build the scientific analysis first. The example uses random-effects
meta-regression because its formula contains subject covariates. The
intercept-only `meta:re` reducer does not consume a formula; use
`meta:re_reg` when the expected response depends on group, age, site, or other
subject variables.

```{r build-analysis}
analysis <- as_plan(subject_maps) |>
  reduce(
    method = "meta:re_reg",
    formula = ~ group + age + site
  ) |>
  posthoc("fdr:bh")
```

Here, `grouppatient` is the patient-minus-control coefficient implied by the
factor coding in `~ group + age + site`. It is an **estimand**: the particular
coefficient, or linear combination of coefficients, whose group statistic is
being tested for sensitivity. Surprise depends on the fitted response model;
influence additionally depends on the estimand. A subject can therefore be
influential for one coefficient and not another.

`examine_group()` locates that reducer, executes only its subject-level prefix,
and inherits its formula, options, and weighting rules. You name the group
estimand once. Quality variables are descriptive unless you also give them an
absolute direction and threshold in `examination_control()`.

```{r configure-examination}
exam_control <- examination_control(
  review = list(
    quality = list(
      mean_fd = list(direction = "high", threshold = 0.5)
    )
  )
)
```

```{r run-examination}
exam <- examine_group(
  analysis,
  estimands = "grouppatient",
  quality = "mean_fd",
  control = exam_control
)

summary(exam)
```

```{r validate-examination, include=FALSE}
stopifnot(
  inherits(exam, "gds_examination"),
  identical(exam$cohort$model, "meta:re_reg"),
  identical(unique(exam$estimand_data$estimand), "grouppatient"),
  all(exam$subject_data$review_status %in% c("none", "review", "insufficient")),
  identical(exam$subject_data$subject[exam$subject_data$retained], exam$config$retained_subjects),
  is.finite(exam$embedding$captured_energy),
  exam$embedding$captured_energy > 0,
  exam$embedding$captured_energy <= 1
)
```

## Why are particular subjects highlighted?

The review queue is created before plotting. A subject receives
`review_status = "review"` when at least one complete absolute criterion is
met anywhere in the requested contrasts and estimands:

| Signal | Complete criterion |
|---|---|
| Surprise | energy at least `r exam$config$control$review$surprise$energy_threshold` **and** tail extent at least `r exam$config$control$review$surprise$tail_threshold` **and** stability at least `r exam$config$control$review$min_stability` |
| Influence | energy at least `r exam$config$control$review$influence$energy_threshold` **and** maximum absolute deletion statistic at least `r exam$config$control$review$influence$max_abs_threshold` **and** stability at least `r exam$config$control$review$min_stability` |
| Quality | any configured absolute quality threshold; here, mean FD at least `r exam$config$control$review$quality$mean_fd$threshold` |

`review_priority` is different: it is the largest cohort percentile across
quality, surprise, and influence. It ranks cases for limited map retention, but
a high percentile alone does not create a review flag. Among criteria that do
trigger, `review_source` records the source with the highest percentile and
`review_reason` records the corresponding component.

```{r review-table}
review <- exam$subject_data[
  exam$subject_data$review_status == "review",
  c("subject", "review_source", "review_priority", "review_reason")
]
review
```

```{r build-review-evidence, include=FALSE}
surprise_gate <- exam$config$control$review$surprise
min_stability <- exam$config$control$review$min_stability
surprise_case <- review$subject[review$review_source == "surprise"][1]
quality_case <- review$subject[review$review_source == "quality"][1]
surprise_rows <- exam$contrast_data[
  exam$contrast_data$subject == surprise_case,
  , drop = FALSE
]
surprise_hit <- with(
  surprise_rows,
  surprise_energy >= surprise_gate$energy_threshold &
    tail_extent >= surprise_gate$tail_threshold &
    surprise_stability >= min_stability
)
surprise_row <- surprise_rows[which(surprise_hit)[1], , drop = FALSE]
quality_value <- exam$subject_data$mean_fd[
  match(quality_case, exam$subject_data$subject)
]
quality_threshold <- exam$config$control$review$quality$mean_fd$threshold
review_evidence <- rbind(
  data.frame(
    subject = surprise_case,
    source = "model surprise",
    component = surprise_row$contrast,
    observed = sprintf(
      "energy %.2f; tail %.3f; stability %.2f",
      surprise_row$surprise_energy,
      surprise_row$tail_extent,
      surprise_row$surprise_stability
    ),
    required = sprintf(
      "energy >= %.1f; tail >= %.2f; stability >= %.1f",
      surprise_gate$energy_threshold,
      surprise_gate$tail_threshold,
      min_stability
    ),
    stringsAsFactors = FALSE
  ),
  data.frame(
    subject = quality_case,
    source = "data quality",
    component = "mean_fd",
    observed = sprintf("mean_fd %.2f", quality_value),
    required = sprintf("mean_fd >= %.2f", quality_threshold),
    stringsAsFactors = FALSE
  )
)
```

```{r review-evidence, echo=FALSE}
knitr::kable(review_evidence, align = "lllll")
```

```{r validate-review, include=FALSE}
stopifnot(
  nrow(review) > 0,
  any(review$review_source == "surprise"),
  any(review$review_source == "quality"),
  nrow(surprise_row) == 1L,
  is.finite(quality_value),
  quality_value >= quality_threshold
)
```

In this example, `r surprise_case` is highlighted because its reversed
`faces>places` pattern passes all three surprise requirements; its negative
map gain adds useful context but is not itself a review gate. `r quality_case`
is highlighted because its `mean_fd` crosses the declared quality threshold,
even though its location in the action plane is unremarkable.

## How do you read the action plane?

Each point represents one subject for one contrast and one estimand. The
horizontal coordinate is surprise energy for that subject and contrast. The
vertical coordinate is influence energy for that subject, contrast, and
estimand. The plot below selects one contrast so its interpretation is direct;
omit `contrast` to facet over all contrasts.

The dashed lines show only the **energy components** of the surprise and
influence criteria. They are visual reference lines, not complete decision
boundaries: tail extent, maximum absolute deletion, and stability are also
required. Yellow fill marks membership in the cohort-wide review queue, a red
outline marks a quality-sourced review, shape represents site, and opacity
records the weaker of the surprise and influence split-stability estimates.

Labels are also cohort-wide. Once a subject enters the review queue because of
any contrast, estimand, or configured quality criterion, its label is shown in
every displayed facet. Consequently, a labeled point need not cross either
dashed line in the particular panel where it appears.

```{r action-plane, fig.width=9, fig.height=5, fig.cap="Labels mark the cohort-wide review queue. A label may reflect another component, such as the configured quality criterion, rather than the point's position relative to the energy reference lines."}
plot(
  exam,
  contrast = "faces>places",
  estimand = "grouppatient",
  group_by = "site"
)
```

A point on the right has a large model-adjusted predictive residual pattern. A
high point changes the patient-control group statistic when deleted. Neither
position supplies a diagnosis or a removal rule. The review table, not visual
quadrant membership, is the authoritative explanation of why a subject was
highlighted.

## What does the residual geometry add?

Surprise energy reduces a whole residual map to one magnitude. Two subjects can
have the same energy while deviating in different features or opposite
directions. The residual geometry keeps that pattern information: it embeds
the signed, capped, model-adjusted residual maps in a low-rank coordinate
system.

Nearby points have similar residual patterns; distant points differ in pattern.
The axes are data-derived directions, not named biological variables, and the
view is not a clustering or classification result. It can suggest a stable
subgroup, a site-associated gradient, or an isolated pattern that merits
inspection. The caption reports the fraction of total residual energy captured
by the retained dimensions, and opacity reports coordinate stability under
deterministic feature splits.

```{r residual-geometry, fig.cap="Low-rank geometry of signed, model-adjusted predictive residuals."}
plot(exam, type = "embedding", group_by = "site")
```

The heatmap keeps contrast-specific surprise, signed map gain, and
estimand-specific influence visible without collapsing them into one score.
Surprise and influence are shown as within-cohort ranks; map gain is shown as a
signed deviation from one. It is therefore a navigation view across metrics,
not a common numerical scale for comparing their raw magnitudes.

```{r diagnostic-heatmap, fig.width=9, fig.height=4.5, fig.cap="Subject summaries across contrasts and the requested estimand."}
plot(exam, type = "heatmap")
```

## How do you inspect one retained subject?

The second pass retains maps for the review queue, explicitly requested
subjects, and the configured highest-priority cases. A retained subject is not
necessarily flagged for review: `retain_n` deliberately keeps a bounded set of
high-ranking cases so you can localize near-threshold patterns without changing
the absolute review rule. Choose a reviewed ID to compare the observed map,
leave-one-out expectation, standardized residual, and change in the group
statistic.

```{r choose-subject}
selected_subject <- review$subject[1]
selected_subject
```

```{r selected-subject, fig.width=9, fig.height=7, fig.cap="Linked profiles for one retained subject."}
plot(exam, subject = selected_subject)
```

For a random-effects model, screening every subject at every sample with a full
heterogeneity refit would be expensive and would make the ranking depend on
different re-estimated variance surfaces. The first pass therefore uses
analytic deletion with the full-data heterogeneity estimate held fixed. This is
the `tau2_fixed_full` screening mode.

Retained subjects then receive an exact refit that re-estimates heterogeneity
(`tau2_refit_exact`). The `mode` and `ranking_stage` columns keep screening and
localization distinguishable. Exact refits do not recursively change the
review queue; freezing selection before the second pass avoids a circular
procedure in which refinement changes which cases are refined.

```{r sensitivity-modes}
unique(exam$estimand_data[c("mode", "ranking_stage")])
```

```{r validate-random-effects, include=FALSE}
stopifnot(
  "tau2_fixed_full" %in% exam$estimand_data$mode,
  "tau2_refit_exact" %in% exam$estimand_data$mode,
  all(exam$estimand_data$ranking_stage %in% c("screening", "selected_refit"))
)
```

## What happens to the post-hoc conclusion?

Because the analysis includes BH FDR, the examination recomputes that
threshold-dependent conclusion for retained subjects only. These counts and
maps live under `exam$conclusion`, separate from the continuous influence
metrics. A custom post-hoc method without a case-deletion contract remains
available for the full fit but is reported as unavailable for deleted cases.

```{r conclusion-summary}
head(exam$conclusion$results)
```

```{r validate-conclusion, include=FALSE}
stopifnot(
  inherits(exam$conclusion$full_maps, "gds"),
  inherits(exam$conclusion$deleted_maps, "gds"),
  all(exam$conclusion$results$status == "available"),
  all(exam$conclusion$results$mode == "tau2_refit_exact")
)
```

## How do you share the examination?

`write_report()` creates one HTML file with inline CSS and SVG. The action
plane and geometry link to retained-subject sections. The file has no external
font, image, script, or stylesheet dependency.

```{r write-report}
report_file <- file.path(tempdir(), "fmrigds-group-examination-report.html")
write_report(exam, report_file, overwrite = TRUE)
report_file
```

The examination and the population fit remain separate branches. Compute the
fit when you are ready to interpret the group result:

```{r compute-fit}
fit <- compute(analysis)
names(assays(fit))
```

If only effect maps are available, use `ols:voxelwise`. The examination labels
that route as effect-only and does not interpret synthetic unit variance as
measured precision. For configuration details, see `?examine_group` and
`?examination_control`.

The intended endpoint is an auditable review, not automatic exclusion. For a
highlighted subject, inspect the recorded reason, its contrast- or
estimand-specific row, the retained spatial profiles, and the selected-deletion
conclusion. Then document any source-data check or sensitivity analysis outside
the original fit. The examination preserves that separation by never deleting
a subject or rewriting the analysis plan.
