---
title: "Examine subjects under a group model"
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: lapis
  preset: study
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts
vignette: |
  %\VignetteIndexEntry{Examine subjects under a group model}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("albersdown", quietly = TRUE)) {
  if (requireNamespace("systemfonts", quietly = TRUE)) try(albersdown::albers_register_fonts(), silent = TRUE)
  if (requireNamespace("ggplot2", quietly = TRUE)) try(ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset)), silent = TRUE)
}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
library(fmrigds)
set.seed(20260805)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

You have fitted a group model, but its average alone cannot tell you whether a
participant is unexpected for their age, group, or site, or whether that
participant materially changes the scientific result. These are different
questions. A noisy estimate can look unusual and carry little weight; a precise,
otherwise conventional estimate can have substantial influence.

`examine_group()` evaluates those questions against the model already present
in an fmrigds plan. It returns subject summaries, group diagnostic maps, a
model-adjusted residual geometry, retained-subject maps, sensitivity results,
and explicit availability states. It never changes the source analysis.

## What data do you need?

The most informative examination starts from subject effect estimates and
genuine first-level variances. Subject covariates and first-level quality
metrics belong in `col_data`. The example below has two task contrasts, a group
effect, an age term, two acquisition sites, one unusually precise subject, and
one subject whose first contrast is reversed.

```{r make-example, include=FALSE}
n_subject <- 16L
n_sample <- 72L
subject_id <- sprintf("sub-%03d", seq_len(n_subject))
group <- factor(rep(c("control", "patient"), each = n_subject / 2))
site <- factor(rep(c("site-a", "site-b"), length.out = n_subject))
age <- seq(21, 62, length.out = n_subject)
mean_fd <- c(runif(n_subject - 1L, 0.08, 0.22), 0.56)
participants <- data.frame(group, age, site, mean_fd, row.names = subject_id)

feature <- seq(0, 2 * pi, length.out = n_sample)
patterns <- cbind(sin(feature), cos(feature))
beta <- array(NA_real_, c(n_sample, n_subject, 2L))
for (i in seq_len(n_subject)) {
  group_shift <- if (group[i] == "patient") 0.35 else 0
  age_shift <- 0.006 * (age[i] - mean(age))
  site_shift <- if (site[i] == "site-b") 0.12 else 0
  beta[, i, ] <- patterns + group_shift + age_shift + site_shift +
    matrix(rnorm(n_sample * 2L, sd = 0.12), n_sample, 2L)
}
beta[, 15L, 1L] <- -patterns[, 1L] + rnorm(n_sample, sd = 0.05)
variance <- array(0.08, dim(beta))
variance[, 8L, ] <- 0.006

subject_maps <- new_gds(
  assays = list(beta = beta, var = variance),
  space = space_sample_labels(sprintf("feature-%03d", seq_len(n_sample))),
  subjects = subject_id,
  contrasts = c("faces>places", "tools>faces"),
  col_data = participants
)
```

```{r inspect-input}
subject_maps
head(col_data(subject_maps))
```

## How do you run the examination?

Build the scientific analysis first. The example uses random-effects
meta-regression because its formula contains subject covariates. The
intercept-only `meta:re` reducer does not consume a formula; use
`meta:re_reg` when the expected response depends on group, age, site, or other
subject variables.

```{r build-analysis}
analysis <- as_plan(subject_maps) |>
  reduce(
    method = "meta:re_reg",
    formula = ~ group + age + site
  ) |>
  posthoc("fdr:bh")
```

`examine_group()` locates that reducer, executes only its subject-level prefix,
and inherits its formula, options, and weighting rules. You name the group
estimand once. Quality variables are descriptive unless you also give them an
absolute direction and threshold in `examination_control()`.

```{r run-examination}
exam <- examine_group(
  analysis,
  estimands = "grouppatient",
  quality = "mean_fd",
  control = examination_control(
    review = list(
      quality = list(
        mean_fd = list(direction = "high", threshold = 0.5)
      )
    )
  )
)

summary(exam)
```

```{r validate-examination, include=FALSE}
stopifnot(
  inherits(exam, "gds_examination"),
  identical(exam$cohort$model, "meta:re_reg"),
  identical(unique(exam$estimand_data$estimand), "grouppatient"),
  all(exam$subject_data$review_status %in% c("none", "review", "insufficient")),
  identical(exam$subject_data$subject[exam$subject_data$retained], exam$config$retained_subjects),
  is.finite(exam$embedding$captured_energy),
  exam$embedding$captured_energy > 0,
  exam$embedding$captured_energy <= 1
)
```

## How do you read the action plane?

The default plot keeps surprise and influence on separate axes. The dashed
lines show the configured absolute review gates. Point opacity records the
split-feature stability estimate. Only subjects that meet a gate receive a
label; all subjects remain in the result.

```{r action-plane, fig.cap="The action plane separates model surprise from influence on the patient-control estimand."}
plot(exam, group_by = "site")
```

A point on the right has a large externally predictive residual pattern after
adjustment for group, age, and site. A high point changes the requested group
statistic when deleted. Neither position supplies a diagnosis or a removal
rule. Use `exam$subject_data$review_reason` and the contrast- and
estimand-specific tables to see which component set the review status.

```{r review-table}
review <- exam$subject_data[
  exam$subject_data$review_status != "none",
  c("subject", "review_source", "review_priority", "review_reason")
]
review
```

```{r validate-review, include=FALSE}
stopifnot(
  nrow(review) > 0,
  any(review$review_source == "surprise"),
  any(review$review_source == "quality")
)
```

## What does the residual geometry add?

The embedding summarizes model-adjusted residual patterns. It can reveal a
stable subgroup, a site-associated gradient, or an isolated pattern, but it is
not a clustering result. The caption reports how much residual energy the
retained dimensions capture, and opacity reports stability under deterministic
feature splits.

```{r residual-geometry, fig.cap="Low-rank geometry of signed, model-adjusted predictive residuals."}
plot(exam, type = "embedding", group_by = "site")
```

The heatmap keeps contrast-specific surprise, signed map gain, and
estimand-specific influence visible without collapsing them into one score.

```{r diagnostic-heatmap, fig.width=9, fig.height=4.5, fig.cap="Subject summaries across contrasts and the requested estimand."}
plot(exam, type = "heatmap")
```

## How do you inspect one retained subject?

The second pass retains maps only for the review queue, explicitly requested
subjects, and the configured highest-priority cases. Choose one of those IDs to
compare the observed map, leave-one-out expectation, standardized residual,
and change in the group statistic.

```{r choose-subject}
selected_subject <- exam$config$retained_subjects[1]
selected_subject
```

```{r selected-subject, fig.width=9, fig.height=7, fig.cap="Linked profiles for one retained subject."}
plot(exam, subject = selected_subject)
```

For a random-effects model, the opening ranking uses analytic deletion with
the full-data heterogeneity estimate held fixed. Retained subjects also receive
an exact refit that re-estimates heterogeneity. The `mode` and `ranking_stage`
columns keep those results distinguishable; exact refits do not recursively
change the review queue.

```{r sensitivity-modes}
unique(exam$estimand_data[c("mode", "ranking_stage")])
```

```{r validate-random-effects, include=FALSE}
stopifnot(
  "tau2_fixed_full" %in% exam$estimand_data$mode,
  "tau2_refit_exact" %in% exam$estimand_data$mode,
  all(exam$estimand_data$ranking_stage %in% c("screening", "selected_refit"))
)
```

## What happens to the post-hoc conclusion?

Because the analysis includes BH FDR, the examination recomputes that
threshold-dependent conclusion for retained subjects only. These counts and
maps live under `exam$conclusion`, separate from the continuous influence
metrics. A custom post-hoc method without a case-deletion contract remains
available for the full fit but is reported as unavailable for deleted cases.

```{r conclusion-summary}
head(exam$conclusion$results)
```

```{r validate-conclusion, include=FALSE}
stopifnot(
  inherits(exam$conclusion$full_maps, "gds"),
  inherits(exam$conclusion$deleted_maps, "gds"),
  all(exam$conclusion$results$status == "available"),
  all(exam$conclusion$results$mode == "tau2_refit_exact")
)
```

## How do you share the examination?

`write_report()` creates one HTML file with inline CSS and SVG. The action
plane and geometry link to retained-subject sections. The file has no external
font, image, script, or stylesheet dependency.

```{r write-report}
report_file <- file.path(tempdir(), "fmrigds-group-examination-report.html")
write_report(exam, report_file, overwrite = TRUE)
report_file
```

The examination and the population fit remain separate branches. Compute the
fit when you are ready to interpret the group result:

```{r compute-fit}
fit <- compute(analysis)
names(assays(fit))
```

If only effect maps are available, use `ols:voxelwise`. The examination labels
that route as effect-only and does not interpret synthetic unit variance as
measured precision. For configuration details, see `?examine_group` and
`?examination_control`.
