---
title: "Getting Started with fmrigds"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{Getting Started with fmrigds}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE
)
```

## Introduction

**fmrigds** (Group Data Sets for fMRI) provides a unified framework for
group-level analysis of first-level fMRI statistical outputs. Whether your data
lives in CSV files, NIfTI volumes, HDF5 stores, or parcel-based summaries,
fmrigds gives you a single, consistent API to:

- Load data from any supported format
- Build analysis pipelines declaratively (lazy evaluation)
- Derive statistics (t, z, p-values) with correct variance propagation
- Run group-level analyses (fixed-effects, random-effects, meta-regression)
- Export results to multiple formats

The key insight is that fMRI group analysis shares common patterns regardless
of whether you're working with voxels, parcels, or latent components. fmrigds
abstracts these patterns into a clean, composable interface.

## Core Concepts

### The GDS Object

A **GDS** (Group Data Set) is a container that holds:

- **Assays**: Named statistical maps (beta, variance, t, etc.)
- **Space**: The spatial representation (voxels, parcels, basis components)
- **Subjects**: Who contributed data
- **Contrasts**: What conditions are being compared
- **Metadata**: Subject-level covariates and provenance information

### Lazy Evaluation

fmrigds uses lazy evaluation: you build a **plan** describing what you want to
do, then **compute** it when ready. This has several advantages:

1. **Efficiency**: Operations are optimized and streamed for large datasets
2. **Clarity**: You can inspect and modify the plan before execution
3. **Reproducibility**: Plans are self-documenting analysis pipelines

```r
# This doesn't load data yet - it creates a plan
plan <- gds("my_data.csv") |>
  derive("t") |>
  reduce(method = "fixed")

# Now execute the plan
result <- compute(plan)
```

## Quick Start: Tabular Data

Let's walk through a complete example using tabular (CSV) data, the simplest
format to get started with.

### Step 1: Prepare Your Data

fmrigds expects tabular data with these columns:

- `sample`: Spatial unit identifier (e.g., "ROI_1", "voxel_12345")
- `subject`: Subject identifier (e.g., "sub-01")
- `contrast`: Condition label (e.g., "faces_vs_places")
- `beta`: Effect estimate
- `var` or `se`: Variance or standard error

```r
library(fmrigds)

# Create example data
tmp <- tempfile(fileext = ".csv")
writeLines(c(
  "sample,subject,contrast,beta,var",
  "ROI_1,sub-01,cope1,0.5,0.04",
  "ROI_2,sub-01,cope1,0.8,0.09",
  "ROI_1,sub-02,cope1,1.0,0.04",
  "ROI_2,sub-02,cope1,1.2,0.09",
  "ROI_1,sub-03,cope1,0.7,0.04",
  "ROI_2,sub-03,cope1,0.9,0.09"
), tmp)
```

### Step 2: Create a Plan

Point `gds()` at your data file. The format is auto-detected:

```r
plan <- gds(tmp)
plan
#> GDS Plan
#>   Source: tabular (CSV)
#>   Space: sample_labels
#>   Operations: (none pending)
```

### Step 3: Compute and Explore

Execute the plan to get a realized GDS object:

```r
g <- compute(plan)

# What assays do we have?
names(assays(g))
#> [1] "beta" "var"

# What's the spatial layout?
space(g)
#> <space_sample_labels>
#>   labels: ROI_1, ROI_2

# Who are our subjects?
subjects(g)
#> [1] "sub-01" "sub-02" "sub-03"

# What contrasts?
contrasts(g)
#> [1] "cope1"
```

## Deriving Statistics

The `derive()` verb computes new assays from existing ones. fmrigds knows the
statistical relationships between common assay types:

```r
# Derive t-statistics from beta and variance
plan_with_t <- plan |> derive("t")
g2 <- compute(plan_with_t)

names(assays(g2))
#> [1] "beta" "var" "t"

# Derive multiple statistics at once
plan_full <- plan |> derive(c("t", "z", "p"))
g3 <- compute(plan_full)

names(assays(g3))
#> [1] "beta" "var" "t" "z" "p"
```

Available derivations include:

| Target | Requirements | Description |
|--------|--------------|-------------|
| `t` | beta, var | t-statistic |
| `z` | t (or beta, var) | z-score |
| `p` | z (or t) | p-value |
| `se` | var | Standard error |
| `var` | se | Variance |

## Group-Level Analysis with Reducers

The `reduce()` verb combines data across subjects into group-level summaries.

### Fixed-Effects Meta-Analysis

Assumes a common true effect across subjects, weighted by precision:

```r
plan_fe <- plan |> reduce(method = "fixed")
g_fe <- compute(plan_fe)

# Group-level results
names(assays(g_fe))
#> [1] "beta" "var" "z" "p" "Q" "I2"

subjects(g_fe)
#> [1] "meta"

# Access the group effect
assay(g_fe, "beta")
```

### Random-Effects Meta-Analysis

Accounts for between-subject heterogeneity using DerSimonian-Laird:

```r
plan_re <- plan |> reduce(method = "random")
g_re <- compute(plan_re)

# Includes tau2 (between-subject variance)
names(assays(g_re))
#> [1] "beta" "var" "z" "p" "Q" "I2" "tau2"
```

### Meta-Regression

Include subject-level covariates in the analysis:

```r
# Define covariates (row names must match subject IDs)
covariates <- data.frame(
  age = c(25, 32, 28),
  group = c("control", "patient", "control"),
  row.names = c("sub-01", "sub-02", "sub-03")
)

# Attach covariates and run regression
plan_reg <- plan |>
  with_col_data(covariates) |>
  reduce(method = "meta:fe_reg", formula = ~ age + group)

g_reg <- compute(plan_reg)

# Coefficients for each predictor become separate assays
names(assays(g_reg))
#> [1] "coef:(Intercept)" "coef:age" "coef:grouppatient"
#> [4] "se_coef:(Intercept)" "se_coef:age" "se_coef:grouppatient"
```

## Writing Results

Use `write_out()` to save results in various formats:

```r
# Save to HDF5 (preserves all metadata and provenance)
plan_h5 <- plan |>
  reduce(method = "fixed") |>
  write_out("group_results.h5", format = "h5")
compute(plan_h5)

# Save to CSV (for easy import into other tools)
plan_csv <- plan |>
  reduce(method = "fixed") |>
  write_out("group_results.csv", format = "csv")
compute(plan_csv)
```

## Working with NIfTI Files

For voxelwise analyses, fmrigds can read NIfTI statistical maps:

```r
# Single file
plan <- gds("cope1.nii.gz")

# Multiple subjects with beta and variance maps
# Use image_catalog for file discovery (see below)
```

### File Discovery with image_catalog

When you have many NIfTI files across subjects, `image_catalog()` helps
organize them:

```r
# Discover files matching a pattern
catalog <- image_catalog(
  root = "study_dir",
  pattern = "sub-*/stats/*.nii.gz",
  path_regex = "sub-(?<subject>[^/]+)"
)

# Map files to assay types
catalog <- catalog |>
  map_assays(
    beta = "cope[0-9]+\\.nii",
    se = "varcope[0-9]+\\.nii"
  )

# Convert to GDS plan
plan <- as_gds(catalog, mask = "group_mask.nii.gz")
```

## Post-hoc Corrections

Apply multiple comparison corrections after group analysis:

```r
# Benjamini-Hochberg FDR
plan_fdr <- plan |>
  reduce(method = "fixed") |>
  posthoc("fdr:bh")

g_fdr <- compute(plan_fdr)
assay(g_fdr, "q")  # FDR-adjusted q-values
```

Available methods:

- `"fdr:bh"` - Benjamini-Hochberg
- `"fdr:by"` - Benjamini-Yekutieli

## Inspecting Plans

Before computing, you can examine what operations are pending:

```r
plan <- gds(tmp) |>
  derive("t") |>
  reduce(method = "fixed") |>
  posthoc("fdr:bh")

# Human-readable summary
explain(plan)

# Detailed operation list
explain_plan(plan)
```

## Validation

Check that your data and plans are well-formed:

```r
# Validate a GDS object
g <- compute(plan)
validate(g)

# Validate a catalog before conversion
report <- validate(catalog)
print(report)
```

## Summary of Key Functions

| Function | Purpose |
|----------|---------|
| `gds()` | Create a lazy plan from any supported source |
| `compute()` | Execute a plan and return results |
| `derive()` | Add derived statistics (t, z, p, etc.) |
| `reduce()` | Group-level analysis (fixed, random, regression) |
| `posthoc()` | Multiple comparison correction |
| `write_out()` | Save results to file |
| `with_col_data()` | Attach subject-level covariates |
| `assays()` / `assay()` | Access statistical maps |
| `subjects()` | List subjects in the data |
| `contrasts()` | List contrasts in the data |
| `space()` | Get spatial representation info |
| `explain()` | Human-readable summary |
| `validate()` | Check data integrity |

## Next Steps

- **Spatial FDR**: See `vignette("as-plan-and-spatial-fdr")` for spatially-aware
  multiple comparison correction
- **fmristore Integration**: See `vignette("fmristore-ingestion")` for working
  with fmristore HDF5 files
- **Technical Details**: See `TECHNICAL_SPECIFICATION.md` for the full design
  document

## Session Info

```{r session, eval=TRUE}
sessionInfo()
```
