---
title: Repeated-measures mixed models
output:
  rmarkdown::html_vignette:
    toc: yes
    toc_depth: 2.0
    css: albers.css
    includes:
      in_header: albers-header.html
params:
  family: red
  preset: homage
resource_files:
- albers.css
- albers.js
- albers-header.html
- fonts

vignette: |
  %\VignetteIndexEntry{Repeated-measures mixed models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
if (requireNamespace("ragg", quietly = TRUE)) knitr::opts_chunk$set(dev = "ragg_png")
if (requireNamespace("systemfonts", quietly = TRUE)) albersdown::albers_register_fonts()
if (requireNamespace("ggplot2", quietly = TRUE) && requireNamespace("albersdown", quietly = TRUE)) ggplot2::theme_set(albersdown::theme_albers(family = params$family, preset = params$preset))
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
library(fmrigds)
set.seed(1)
```

```{r albers-classes, echo=FALSE, results='asis'}
cat(sprintf(
  paste0(
    '<script>document.addEventListener("DOMContentLoaded",function(){',
    'document.body.classList.remove("palette-red","palette-lapis","palette-ochre","palette-teal","palette-green","palette-violet","preset-homage","preset-study","preset-structural","preset-adobe","preset-midnight");',
    'document.body.classList.add("palette-%s","preset-%s");',
    '});</script>'
  ),
  params$family,
  params$preset
))
```

Repeated-measures fMRI analyses are often statistically simple but operationally
awkward. You have one effect estimate per ROI or voxel, per subject, per visit
or condition, and you want a group model that respects the repeated structure
without hand-writing a loop over `lmer()` calls. `fmrigds` gives you a narrow,
fast path for that case.

This vignette shows the supported contract: one grouping factor, Gaussian
responses, shared observation layout across samples, and either a random
intercept or a random intercept plus one within-subject slope. By the end, you
will have one model-level GDS containing fixed-effect estimates and
variance-component assays that you can inspect or write out like any other
group result.

## What does the input look like?

The easiest route is a stacked long table with one row per sample, subject, and
repeated-measure level. Here we build a small ROI example with three visits:

```{r make-long-table}
subjects <- paste0("sub-", sprintf("%02d", 1:6))
visits <- c("pre", "mid", "post")
time <- c(-1, 0, 1)

rm_df <- expand.grid(
  sample = c("ROI_1", "ROI_2"),
  subject = subjects,
  contrast = visits,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)

subject_intercept <- c(-0.25, -0.10, 0.00, 0.10, 0.20, 0.30)
names(subject_intercept) <- subjects
subject_slope <- c(-0.05, -0.02, 0.00, 0.01, 0.03, 0.05)
names(subject_slope) <- subjects
sample_shift <- c(ROI_1 = 0.00, ROI_2 = 0.35)
sample_slope <- c(ROI_1 = 0.45, ROI_2 = 0.70)

rm_df$time <- time[match(rm_df$contrast, visits)]
rm_df$beta <- 0.9 +
  sample_shift[rm_df$sample] +
  subject_intercept[rm_df$subject] +
  (sample_slope[rm_df$sample] + subject_slope[rm_df$subject]) * rm_df$time
rm_df$var <- 0.04

knitr::kable(head(rm_df), row.names = FALSE)
```

The `time` column is constant within each visit, so it can be collapsed into
contrast-level metadata during ingestion.

## What is the fastest end-to-end path?

Use `gds(..., contrast_data_cols = "time")` to put the repeated-measure
metadata on the contrast axis, then call `reduce()` with one of the restricted
LMM reducers:

```{r quick-win}
g_rm <- gds(rm_df, contrast_data_cols = "time") |>
  reduce(
    method = "lmm:ri_slope1",
    formula = ~ time,
    options = list(
      slope = "time",
      covariance = "diag",
      fit = "REML",
      theta_mode = "voxelwise"
    )
  ) |>
  compute()
```

```{r quick-win-checks, include = FALSE}
lmm_summary <- data.frame(
  sample = sample_labels(g_rm),
  intercept = as.numeric(assay(g_rm, "coef:(Intercept)")),
  time = as.numeric(assay(g_rm, "coef:time")),
  vc_intercept = as.numeric(assay(g_rm, "vc_intercept")),
  vc_slope = as.numeric(assay(g_rm, "vc_slope")),
  lambda_slope = as.numeric(assay(g_rm, "lambda_slope")),
  stringsAsFactors = FALSE
)

lmm_summary_display <- transform(
  lmm_summary,
  intercept = round(intercept, 3),
  time = round(time, 3),
  vc_intercept = round(vc_intercept, 3),
  vc_slope = round(vc_slope, 3),
  lambda_slope = round(lambda_slope, 3)
)

stopifnot(
  identical(subjects(g_rm), "meta"),
  identical(contrasts(g_rm), "model"),
  all(is.finite(lmm_summary$time)),
  all(lmm_summary$vc_slope > 0),
  all(lmm_summary$lambda_slope > 0)
)
```

The result is still a GDS, but the subject and contrast axes have collapsed to
the model level. A compact summary is usually easier to read than printing
individual assay arrays:

```{r inspect-summary}
knitr::kable(lmm_summary_display, row.names = FALSE)
```

Fixed-effect outputs still follow the same naming convention as the regression
reducers (`coef:<term>`, `se_coef:<term>`, `t_coef:<term>`, `p_coef:<term>`),
while variance components are exposed as ordinary assays such as
`vc_intercept`, `vc_slope`, and `lambda_slope`.

## When should you use pooled vs voxelwise theta?

`theta_mode = "pooled"` estimates one shared variance structure across all
samples. `theta_mode = "voxelwise"` estimates variance parameters separately
for each sample. Pooled mode reuses one variance structure, so it is typically
faster; voxelwise mode is more flexible.

```{r pooled-vs-voxelwise}
g_pooled <- gds(rm_df, contrast_data_cols = "time") |>
  reduce(
    method = "lmm:ri_slope1",
    formula = ~ time,
    options = list(
      slope = "time",
      covariance = "diag",
      fit = "REML",
      theta_mode = "pooled"
    )
  ) |>
  compute()

theta_compare <- data.frame(
  sample = sample_labels(g_rm),
  pooled_lambda = round(as.numeric(assay(g_pooled, "lambda_slope")), 3),
  voxelwise_lambda = round(as.numeric(assay(g_rm, "lambda_slope")), 3),
  stringsAsFactors = FALSE
)

knitr::kable(theta_compare, row.names = FALSE)
```

```{r pooled-vs-voxelwise-checks, include = FALSE}
bench_subjects <- paste0("sub-", sprintf("%02d", 1:10))
bench_visits <- c("pre", "mid", "post")
bench_time <- c(-1, 0, 1)
bench_samples <- paste0("ROI_", sprintf("%02d", 1:24))

bench_df <- expand.grid(
  sample = bench_samples,
  subject = bench_subjects,
  contrast = bench_visits,
  KEEP.OUT.ATTRS = FALSE,
  stringsAsFactors = FALSE
)

bench_subject_intercept <- seq(-0.25, 0.25, length.out = length(bench_subjects))
names(bench_subject_intercept) <- bench_subjects
bench_subject_slope <- seq(-0.04, 0.04, length.out = length(bench_subjects))
names(bench_subject_slope) <- bench_subjects
bench_sample_shift <- seq(0.00, 0.45, length.out = length(bench_samples))
names(bench_sample_shift) <- bench_samples
bench_sample_slope <- seq(0.35, 0.85, length.out = length(bench_samples))
names(bench_sample_slope) <- bench_samples

bench_df$time <- bench_time[match(bench_df$contrast, bench_visits)]
bench_df$beta <- 0.8 +
  bench_sample_shift[bench_df$sample] +
  bench_subject_intercept[bench_df$subject] +
  (bench_sample_slope[bench_df$sample] + bench_subject_slope[bench_df$subject]) *
    bench_df$time
bench_df$var <- 0.04

bench_pooled_plan <- gds(bench_df, contrast_data_cols = "time") |>
  reduce(
    method = "lmm:ri_slope1",
    formula = ~ time,
    options = list(
      slope = "time",
      covariance = "diag",
      fit = "REML",
      theta_mode = "pooled"
    )
  )

bench_voxel_plan <- gds(bench_df, contrast_data_cols = "time") |>
  reduce(
    method = "lmm:ri_slope1",
    formula = ~ time,
    options = list(
      slope = "time",
      covariance = "diag",
      fit = "REML",
      theta_mode = "voxelwise"
    )
  )

bench_times <- data.frame(
  theta_mode = c("pooled", "voxelwise"),
  elapsed = c(
    median(replicate(3, unname(system.time(compute(bench_pooled_plan))[["elapsed"]]))),
    median(replicate(3, unname(system.time(compute(bench_voxel_plan))[["elapsed"]])))
  ),
  stringsAsFactors = FALSE
)

stopifnot(
  length(unique(as.numeric(assay(g_pooled, "lambda_slope")))) == 1L,
  all(is.finite(as.numeric(assay(g_rm, "lambda_slope")))),
  bench_times$elapsed[1] < bench_times$elapsed[2]
)
```

```{r pooled-vs-voxelwise-timing}
knitr::kable(transform(bench_times, elapsed = round(elapsed, 3)), row.names = FALSE)
```

## What if your repeated-measure metadata is already separate?

If your `time`, visit order, or other repeated-measure annotations already live
outside the long table, attach them explicitly with `with_contrast_data()`.

```{r attach-contrast-data}
contrast_df <- data.frame(time = time, row.names = visits)

g_attached <- gds(rm_df[c("sample", "subject", "contrast", "beta", "var")]) |>
  with_contrast_data(contrast_df) |>
  reduce(
    method = "lmm:ri_slope1",
    formula = ~ time,
    options = list(
      slope = "time",
      covariance = "diag",
      fit = "REML",
      theta_mode = "pooled"
    )
  ) |>
  compute()

knitr::kable(
  data.frame(
    sample = sample_labels(g_attached),
    time = round(as.numeric(assay(g_attached, "coef:time")), 3),
    stringsAsFactors = FALSE
  ),
  row.names = FALSE
)
```

```{r attach-contrast-data-checks, include = FALSE}
stopifnot(
  max(abs(as.numeric(assay(g_attached, "coef:time")) -
            as.numeric(assay(g_pooled, "coef:time")))) < 1e-10,
  max(abs(as.numeric(assay(g_attached, "lambda_slope")) -
            as.numeric(assay(g_pooled, "lambda_slope")))) < 1e-10
)
```

If you only need a random intercept, use `method = "lmm:ri"` instead:

```{r random-intercept}
g_ri <- gds(rm_df, contrast_data_cols = "time") |>
  reduce(
    method = "lmm:ri",
    formula = ~ time,
    options = list(fit = "REML", theta_mode = "pooled")
  ) |>
  compute()

knitr::kable(
  data.frame(
    sample = sample_labels(g_ri),
    time = round(as.numeric(assay(g_ri, "coef:time")), 3),
    lambda = round(as.numeric(assay(g_ri, "lambda")), 3),
    stringsAsFactors = FALSE
  ),
  row.names = FALSE
)
```

## What is supported, and what is intentionally out of scope?

The repeated-measures LMM reducers are deliberately narrower than `lmer()`:

- one grouping factor only
- Gaussian outcomes only
- shared observation layout across all samples
- random intercept only (`"lmm:ri"`) or random intercept plus one slope (`"lmm:ri_slope1"`)
- no crossed random effects
- no general random-effects formula grammar

That restriction is the point: it lets `fmrigds` reuse one design across many
responses and produce a whole-brain or whole-ROI result with one call to
`reduce()`.

## Where do you go next?

- See `?reduce` for the reducer contract and options.
- See `?with_contrast_data` if your repeated-measure metadata is already
  separate from the long table.
- See `inst/examples/fmrigds_lmm_benchmark.R` for a benchmark harness that
  compares pooled and voxelwise theta modes.
