---
title: "Spatial operations: masking, alignment, and mapping"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Spatial operations: masking, alignment, and mapping}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(fmrigds)
set.seed(42)
```

## When do you need this?

The getting-started vignette (`vignette("fmrigds")`) covers the standard
pipeline: load data, derive statistics, reduce across subjects, and correct
for multiple comparisons. That workflow assumes two things:

1. Every subject has data at every spatial location (no missing voxels).
2. All subjects live in the same coordinate system (a common space).

In practice, neither assumption always holds. This vignette introduces three
verbs that handle the messy cases:

- `mask()` --- removes spatial locations that have missing or unreliable data.
- `align()` --- rotates each subject's data into a shared coordinate system.
- `map_to()` --- transforms data from one spatial representation to another
  (e.g., voxels to parcels).

All three are lazy: they add a step to the analysis plan without doing any
computation until you call `compute()`.

## Building a GDS from in-memory data

The examples in this vignette use synthetic data created in R rather than
files on disk. When your data already lives in memory, `as_gds()` turns it
into a GDS object directly.

The key input is a **named list of 3D arrays**. Each array has dimensions
**samples x subjects x contrasts**:

- **Samples** (first dimension): the spatial units---ROIs, voxels, or
  components.
- **Subjects** (second dimension): one slice per participant.
- **Contrasts** (third dimension): one slice per experimental condition.

Here is a small example with 10 ROIs, 4 subjects, and 1 contrast:

```{r as-gds-array}
betas <- array(rnorm(40, mean = 1), dim = c(10, 4, 1))
vars  <- array(runif(40, 0.01, 0.1), dim = c(10, 4, 1))

g <- as_gds(
  list(beta = betas, var = vars),
  space     = space_parcels(paste0("ROI_", 1:10)),
  subjects  = paste0("sub-", sprintf("%02d", 1:4)),
  contrasts = "faces"
)

explain(g)
```

The `space` argument tells fmrigds what the sample axis represents. Here we
use `space_parcels()` because our samples are named ROIs. For voxelwise
data you would use `space_voxel()` with grid dimensions and an affine
matrix; for latent-space data (PCA, ICA), `space_basis()`.

---

## Masking: dealing with missing data

### Why mask?

In whole-brain analyses, some subjects may have signal dropout or
incomplete coverage in certain brain regions. Those missing values show up
as `NA` in the data arrays. If you run a group-level reduction on data
that contains `NA`s, the missing values propagate into the result and
you lose information at locations where most subjects *do* have data.

The `mask()` verb lets you decide, up front, which spatial locations have
enough data to be worth analysing. It removes the rest before any
downstream computation.

### Setting up the example

Let's inject some missing values into our test data to see masking in
action:

```{r inject-na}
# ROI_3 is missing for subject 1; ROI_7 is missing for subject 2
g$assays$beta[3, 1, 1] <- NA
g$assays$var[3, 1, 1]  <- NA
g$assays$beta[7, 2, 1] <- NA
g$assays$var[7, 2, 1]  <- NA
```

Now ROI_3 and ROI_7 each have one subject with missing data.

### Intersection mask (the default)

The simplest policy is `"intersection"`: keep only locations where *every*
subject has valid (non-`NA`) data. This is the strictest rule---if even one
subject is missing, that location is dropped.

```{r mask-default}
g_masked <- mask(g) |> compute()
explain(g_masked)
```

ROI_3 and ROI_7 were dropped because they had at least one `NA`. The
remaining 8 ROIs all have complete data.

### Threshold mask

The intersection rule can be wasteful when you have many subjects---one
missing subject out of 50 would discard the location entirely. The
`"threshold"` rule is more permissive: it keeps a location if the fraction
of non-missing values exceeds a threshold you choose.

```{r mask-threshold}
# Keep ROIs where at least 75% of values are non-missing
g_thresh <- mask(g, MaskPolicy(rule = "threshold", threshold = 0.75)) |>
  compute()

explain(g_thresh)
```

With a 75% threshold, ROI_3 and ROI_7 both survive: each has 3 out of 4
subjects present (75%), which meets the cutoff.

### Other mask rules

Here is the full set of options:

| Rule | Behaviour |
|---|---|
| `"intersection"` | Keep locations where *all* subjects have data (strictest) |
| `"union"` | Keep locations where *any* subject has data (most permissive) |
| `"threshold"` | Keep locations where the fraction of non-missing values >= `threshold` |
| `"custom"` | You provide a function that returns a logical vector |

A custom mask function receives the list of assay arrays and returns one
`TRUE`/`FALSE` per sample. For example, to keep only ROIs with a strong
average signal:

```{r mask-custom}
high_signal <- function(arrays) {
  rowMeans(arrays$beta, na.rm = TRUE) > 0.5
}

g_custom <- mask(g, MaskPolicy(rule = "custom", custom = high_signal)) |>
  compute()

dim(assay(g_custom, "beta"))
```

### Masking in a pipeline

Because `mask()` is lazy, you can slot it into a larger pipeline. Here we
mask first, then run a group analysis with FDR correction---all as a single
plan:

```{r mask-pipeline}
result <- as_plan(g) |>
  mask(MaskPolicy(rule = "intersection")) |>
  reduce(method = "fixed") |>
  posthoc("fdr:bh") |>
  compute()

names(assays(result))
```

The mask runs before the reduction, so the group model only sees complete
cases.

---

## Alignment: bringing subjects into a common space

### Why align?

When you use dimensionality reduction (PCA, ICA) or functional alignment
(hyperalignment) on fMRI data, each subject ends up in their own coordinate
system. Component 1 for subject A is not the same as component 1 for
subject B---the axes are rotated differently.

Before you can average across subjects, you need to apply a per-subject
rotation matrix that maps each subject's data into a shared "group" space.
This is what `align()` does.

### Creating test data in a latent space

Let's create a GDS where each subject has 5 latent components, 2
experimental conditions, and a subject-specific coordinate system:

```{r align-setup}
k <- 5       # number of latent components
n_sub <- 3   # number of subjects
n_con <- 2   # number of contrasts

betas <- array(rnorm(k * n_sub * n_con), dim = c(k, n_sub, n_con))
vars  <- array(runif(k * n_sub * n_con, 0.01, 0.1), dim = c(k, n_sub, n_con))

# Define the native and target spaces
native_space <- space_basis(k, basis_name = "native_pca")
group_space  <- space_basis(k, basis_name = "group_pca")

subjects <- paste0("sub-", sprintf("%02d", 1:n_sub))

g_latent <- as_gds(
  list(beta = betas, var = vars),
  space     = native_space,
  subjects  = subjects,
  contrasts = paste0("cond_", 1:n_con)
)
```

The `space_basis()` constructor says "these samples are latent components."
We have a native space (each subject's own PCA axes) and a target group
space (shared axes).

### Providing per-subject rotation matrices

In a real analysis, rotation matrices come from Procrustes alignment or
hyperalignment. Here we simulate them using random orthogonal matrices:

```{r make-rotations}
random_orthogonal <- function(k) {
  qr.Q(qr(matrix(rnorm(k * k), k, k)))
}

rotations <- setNames(
  lapply(1:n_sub, function(i) random_orthogonal(k)),
  subjects
)

# Each rotation is a k-by-k orthogonal matrix
dim(rotations[[1]])
```

Each matrix is `k x k` (5 x 5 here). Multiplying a subject's data by
their rotation matrix maps them from native coordinates into group
coordinates.

### Bundling rotations into a MapFamily

fmrigds needs to know not just the matrices but also what spaces they
connect and what properties they have. You bundle this information into
a **MapFamily** object. For orthogonal rotations, use the
`OrthogonalFamily()` helper:

```{r orthogonal-family}
alignment <- OrthogonalFamily(
  name                = "procrustes",
  from_space          = native_space,
  to_space            = group_space,
  matrices_by_subject = rotations
)
```

The `orthogonal = TRUE` trait tells fmrigds that these matrices satisfy
`R'R = I`, which simplifies variance propagation---the variance of a
rotated vector is just the rotated variance, with no cross-terms.

### Running the alignment

Now apply the alignment and compute:

```{r align-compute}
g_aligned <- align(g_latent, alignment) |> compute()

explain(g_aligned)
```

After alignment, the space descriptor has changed from `native_pca` to
`group_pca`. Every subject's data has been rotated through their respective
matrix, and variances have been propagated through the rotation
automatically.

### Alignment followed by group analysis

The common workflow is to align first, then reduce across subjects. Because
subjects are now in a common space, the group average is meaningful:

```{r align-reduce}
result <- as_plan(g_latent) |>
  align(alignment) |>
  reduce(method = "fixed") |>
  compute()

explain(result)
```

### Other alignment types

fmrigds provides helpers for different kinds of per-subject maps:

| Helper | Use case | Key property |
|---|---|---|
| `OrthogonalFamily()` | Procrustes, hyperalignment | Preserves distances (orthogonal) |
| `OTFamily()` | Optimal transport plans | Preserves total mass |
| `WarpFamily()` | Deformable spatial warps | General deformations |
| `MapFamily()` | Any linear transform | Fully customisable |

All follow the same pattern: bundle per-subject operators into a family
object, then pass it to `align()`.

---

## Space transformations with map_to()

### Why transform spaces?

Sometimes you want to change the spatial resolution of your data. Common
scenarios:

- **Voxels to parcels**: aggregate thousands of voxels into a smaller set of
  brain regions.
- **Fine parcellation to coarse**: collapse a detailed atlas (e.g., 400
  regions) into broader networks (e.g., 7 networks).
- **Projection**: map data from one coordinate system to another using a
  known linear transformation.

`map_to()` applies a matrix transformation to the sample axis. If the data
has effect estimates and variances, the variances are propagated through the
transformation automatically.

### Setting up a mapping

The transformation matrix has dimensions **target_samples x source_samples**.
Each row defines how one target location is computed from the source
locations. For averaging, each row sums to 1:

```{r map-to-setup}
# 10 ROIs that we want to average into 3 regions:
#   Region "frontal"   = average of ROIs 1-3
#   Region "parietal"  = average of ROIs 4-6
#   Region "occipital" = average of ROIs 7-10
g_clean <- as_gds(
  list(
    beta = array(rnorm(40, mean = 1.5), dim = c(10, 4, 1)),
    var  = array(runif(40, 0.01, 0.05), dim = c(10, 4, 1))
  ),
  space     = space_parcels(paste0("ROI_", 1:10)),
  subjects  = paste0("sub-", sprintf("%02d", 1:4)),
  contrasts = "faces"
)

# Build the averaging matrix (3 target regions x 10 source ROIs)
W <- matrix(0, nrow = 3, ncol = 10)
W[1, 1:3]  <- 1/3   # frontal = mean of ROI 1-3
W[2, 4:6]  <- 1/3   # parietal = mean of ROI 4-6
W[3, 7:10] <- 1/4   # occipital = mean of ROI 7-10

target <- space_parcels(c("frontal", "parietal", "occipital"))
```

### Applying the transformation

Wrap the matrix in a `map_linear()` object that records the source and
target spaces, then pass it to `map_to()`:

```{r map-to-compute}
mapping <- map_linear(
  from_space = space_parcels(paste0("ROI_", 1:10)),
  to_space   = target,
  operator   = W
)

g_regions <- map_to(g_clean, target_space = target, map = mapping) |>
  compute()

explain(g_regions)
assay(g_regions, "beta")
```

Each target region now holds the average of its source ROIs. The variances
have also been propagated: under the default assumption that source
variances are independent, `Var(Wx) = W diag(v) W'`.

### How uncertainty propagation works

When you transform data through a linear map, the variances need to be
transformed too. By default, `map_to()` assumes the source locations have
independent variances (no spatial correlation). If your data has spatial
correlations, you can provide a covariance function:

```{r uncertainty-rules, eval = FALSE}
# Default: independent variances
map_to(g, target, mapping, uncertainty = UncertaintyRule("independent"))

# Correlated variances: provide a function that returns the covariance matrix
map_to(g, target, mapping, uncertainty = UncertaintyRule(
  "cov_provider",
  cov_provider = my_cov_function
))
```

### Combining test statistics without effect estimates

Sometimes your GDS contains only test statistics (z-scores or p-values)
without the underlying effect estimates and variances. In that case, linear
averaging does not apply. Instead, `map_to()` can combine the statistics
using established meta-analytic methods:

- **Stouffer's method**: produces a weighted z-score for each target
  location.
- **Fisher's method**: combines p-values using a chi-squared statistic.

```{r combine-stats}
# A GDS with only z-scores (no beta or var)
z_vals <- array(rnorm(40, mean = 2), dim = c(10, 4, 1))
g_zonly <- as_gds(
  list(z = z_vals),
  space     = space_parcels(paste0("ROI_", 1:10)),
  subjects  = paste0("sub-", sprintf("%02d", 1:4)),
  contrasts = "faces"
)

g_combined <- map_to(
  g_zonly,
  target_space = target,
  map = W,
  combine = "stouffer"
) |> compute()

names(assays(g_combined))
assay(g_combined, "z")
```

The combined z-scores reflect the pooled evidence across the source
locations in each target region.

---

## Putting it all together

These verbs compose into a single pipeline. Here is a workflow that aligns
subjects into a common latent space, then runs a group analysis:

```{r full-pipeline}
k <- 6
n_sub <- 3
betas <- array(rnorm(k * n_sub, mean = 1), dim = c(k, n_sub, 1))
vars  <- array(runif(k * n_sub, 0.01, 0.05), dim = c(k, n_sub, 1))

native <- space_basis(k, "native")
group  <- space_basis(k, "group")
subjects <- paste0("sub-", sprintf("%02d", 1:n_sub))

g_raw <- as_gds(
  list(beta = betas, var = vars),
  space = native, subjects = subjects, contrasts = "task"
)

# Per-subject rotation matrices (simulated)
rots <- setNames(
  lapply(1:n_sub, function(i) {
    qr.Q(qr(matrix(rnorm(k * k), k, k)))
  }),
  subjects
)

fam <- OrthogonalFamily("hyper", native, group, rots)

# The full pipeline: align into group space, then reduce across subjects
result <- as_plan(g_raw) |>
  align(fam) |>
  reduce(method = "fixed") |>
  compute()

explain(result)
```

The pipeline reads left to right: rotate each subject into the group
coordinate system, then pool their data using inverse-variance weighting.

**A note on ordering.** `mask()` changes the number of spatial samples
(it removes locations), while `align()` expects the input dimension to
match the rotation matrix size. If you need both, run `align()` first to
rotate into the common space, *then* `mask()` to remove locations with
missing data, and finally `reduce()`.

---

## Space types at a glance

fmrigds supports five spatial representations. Each has a constructor
function that records the relevant metadata:

| Constructor | What it represents | Typical use |
|---|---|---|
| `space_parcels()` | Named brain regions | ROI or atlas-based analyses |
| `space_voxel()` | 3D volumetric grid | Whole-brain voxelwise analyses |
| `space_surface()` | Cortical triangle mesh | Surface-based analyses |
| `space_basis()` | Latent components | PCA, ICA, hyperalignment |
| `space_sample_labels()` | Generic labels | Tabular/CSV data (the default) |

Here are a few examples:

```{r space-examples}
# A parcels space with four brain regions
sp_parcels <- space_parcels(c("V1", "V2", "FFA", "PPA"))
sp_parcels

# A latent basis space with 20 ICA components
sp_basis <- space_basis(k = 20, basis_name = "ICA")
sp_basis

# A tiny voxel space (3x3x3 grid, 10 active voxels)
sp_voxel <- space_voxel(
  dim         = c(3, 3, 3),
  affine      = diag(4),
  mask_bitmap = array(c(rep(TRUE, 10), rep(FALSE, 17)), dim = c(3, 3, 3)),
  storage     = "packed"
)
sp_voxel
```

The space descriptor travels with the GDS through every operation. After a
`map_to()`, the space updates to reflect the target. After a `mask()`, labels
for removed locations are dropped. You never need to track this manually.

## Next steps

- `vignette("fmrigds")` --- the core pipeline (load, derive, reduce,
  correct)
- `vignette("as-plan-and-spatial-fdr")` --- standard and spatial FDR
  correction
- `vignette("fmristore-ingestion")` --- reading fmristore HDF5 files
- `?align` and `?MapFamily` for alignment details
- `?mask` and `?MaskPolicy` for masking options
- `?map_to` and `?map_linear` for space transformations
